---
title: "PCA project analysis GeomMX copy number alteration "
output: html_notebook
author: Yingxiao Shi(TK)
---

```{r library load, massage = FALSE, warning = FALSE, include = FALSE}
library(NanoStringNCTools)
library(GeomxTools)
library(GeoMxWorkflows)
library(ggthemes)
library(ggiraph)

# library for DEG analysis
library(clusterProfiler)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(enrichplot)
library(knitr)

# library for plotting
library(dplyr)
library(ggforce)
library(ggplot2)
library(scales)

library(reshape2)  # for melt
library(cowplot)
library(umap)
library(Rtsne)
library(DESeq2)
library(tidyverse)
library(pheatmap)
library(ggrepel)
require(gridExtra)
library(repr)
library(readxl)

# for single gene signature analysis
library(GSVA)
library(som)
library(ComplexHeatmap)
library(colorspace)
library(oposSOM)

library(ggstatsplot)
library(Biobase)

library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
library(grid)  # For handling grid graphics objects
library(ggplot2)  # For using ggsave
library(dendextend)
library(seriation)

```

signatures load (really need to organize them)
```{r, signature analysis}
# set the working directory
setwd("/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/Project2_stable_disease_project/gene_signatures/")

# load all the signatures
metabolism_geneSig <- read.csv("/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/Project2_stable_disease_project/gene_signatures/KEGG_metabolism_pathway.csv")[-1]
metabolism_geneSigList <- as.list(metabolism_geneSig)
metabolism_geneSigList <- lapply(metabolism_geneSigList, function(x) x[x != ""])

geneSig <- read.csv("/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/Project2_stable_disease_project/gene_signatures/2018_cell_type_markers.csv", skip = 1, header = T)
cyclin_geneSig <- as.list(geneSig[, c("CELL.CYCLE..G1.S", "CELL.CYCLE..G2.M")])
cyclin_geneSig <- lapply(cyclin_geneSig, function(x) x[x != ""])

MITF_geneSig <- read.csv("/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/Project2_stable_disease_project/gene_signatures/mitf_axl_from_tirosh2016.txt", sep = "\t" )
mitf_geneSig <- as.list(MITF_geneSig[, c("MITF_SIG", "AXL_SIG", "QUIESCENCE")])
mitf_geneSig <- lapply(mitf_geneSig, function(x) x[x != ""])

# now looking at the seven gene signature
seven_geneSig <- read.csv("/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/Project2_stable_disease_project/GeoMX_analysis/seven_tumorsigDF.csv")[-1]
seven_geneSigList <- lapply(as.list(seven_geneSig), function(x) x[x != ""])


# now take the hallmark pathway gene signatures
hallmark_pathway_geneSig <- read.csv("/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/Project2_stable_disease_project/GeoMX_analysis/hallmark_pathway.csv")[-1]
hallmark_pathway_geneSigList <- lapply(as.list(hallmark_pathway_geneSig), function(x) x[x != ""])
#hallmark_pathway_geneSigList[0:1]

cyclin_geneSig["QUIESCENCE"] <- mitf_geneSig["QUIESCENCE"]
#mitf_geneSig <- mit_geneSig.pop("QUIESCENCE")
mitf_geneSig <- mitf_geneSig[c("MITF_SIG", "AXL_SIG")]

Neocroptosis <- read.csv("/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/Project2_stable_disease_project/gene_signatures/GOBP_NECROPTOTIC_SIGNALING_PATHWAY.v2023.1.Hs.tsv", sep = "\t")
Pyroptosis <- read.csv("/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/Project2_stable_disease_project/gene_signatures/REACTOME_PYROPTOSIS.v2023.1.Hs.tsv", sep = "\t")
Ferprotosis <- read.csv("/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/Project2_stable_disease_project/gene_signatures/WP_FERROPTOSIS.v2023.1.Hs.tsv", sep = "\t")

NEOCROPTOSIS_list <- strsplit(Neocroptosis$GOBP_NECROPTOTIC_SIGNALING_PATHWAY[17], ",")[[1]]
PYROPTOSIS_list <- strsplit(Pyroptosis$REACTOME_PYROPTOSIS[17], ",")[[1]]
FERPROTOSIS_list <- strsplit(Ferprotosis$WP_FERROPTOSIS[17], ",")[[1]][-1]
Overall_geneSig <- c(cyclin_geneSig, mitf_geneSig, hallmark_pathway_geneSigList)
Overall_geneSig$NEOCROPTOSIS <- c(NEOCROPTOSIS_list)
Overall_geneSig$FERPROPTOSIS <- c(FERPROTOSIS_list)
Overall_geneSig$PYROPTOSIS <- c(PYROPTOSIS_list)
Overall_geneSig$SENESCENCE <- c("E2F1", "E2F3", "NF1", "PTEN") 

selected_geneSig <- Overall_geneSig[c("CELL.CYCLE..G1.S", "CELL.CYCLE..G2.M", 
                                      "QUIESCENCE", 
                                      "HALLMARK_APOPTOSIS", "NEOCROPTOSIS", "FERPROPTOSIS", "PYROPTOSIS"
                                      # ,"MITF_SIG", "AXL_SIG"
)]

summary(selected_geneSig)
names(selected_geneSig) <- c("G1S_SIG(43)", "G2M_SIG(54)", "QUIESCENCE_SIG(49)", 
                             "APOPTOSIS_SIG(183)", "NECROPTOSIS_SIG(8)", "FERROPTOSIS_SIG(64)", "PYROPTOSIS_SIG(30)"
                             #, "MITF_SIG(100)","AXL_SIG(100)"
)

selected_geneSig$`APOPTOSIS_SIG(183)` <- unique(selected_geneSig$`APOPTOSIS_SIG(183)`)

names(selected_geneSig) <- c("G1S_SIG(43)", "G2M_SIG(54)", "QUIESCENCE_SIG(49)", 
                             "APOPTOSIS_SIG(161)", "NECROPTOSIS_SIG(8)", "FERROPTOSIS_SIG(64)", "PYROPTOSIS_SIG(30)"
                             #, "MITF_SIG(100)","AXL_SIG(100)"
)

# immune gene signatures
# try new immune related pathays
immune_geneSig <- read.csv("/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/Project2_stable_disease_project/GeoMX_analysis/gene_signatures/immune_geneSig.csv")
immune_geneSig_list <- list()

for (sig in unique(immune_geneSig$Gene.signature)[-1]){
  #print(sig)
  if (sig != "") {
    subset_df <- subset(immune_geneSig, Gene.signature == sig)
    #print(subset_df)
    immune_geneSig_list[[sig]] <- subset_df$Gene}}
```

```{r create the tumor signatures}
seven_geneSig <- read.csv("/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/Project2_stable_disease_project/GeoMX_analysis/seven_tumorsigDF.csv")[-1]
seven_geneSigList <- lapply(as.list(seven_geneSig), function(x) x[x != ""])

all_tumorSig <- read_xlsx("/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/Project2_stable_disease_project/gene_signatures/various_melanoma_transcript_programs.xlsx")

proliferationSig <- read.csv("/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/Project2_stable_disease_project/GeoMX_analysis/seven_tumorsigDF.csv")

geneSig <- read.csv("/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/Project2_stable_disease_project/gene_signatures/2018_cell_type_markers.csv", skip = 1, header = T)

cyclin_geneSig <- as.list(geneSig[, c("CELL.CYCLE..G1.S", "CELL.CYCLE..G2.M")])
cyclin_geneSig <- lapply(cyclin_geneSig, function(x) x[x != ""])

MITF_geneSig <- read.csv("/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/Project2_stable_disease_project/gene_signatures/mitf_axl_from_tirosh2016.txt", sep = "\t" )
mitf_geneSig <- as.list(MITF_geneSig[, c("MITF_SIG", "AXL_SIG", "QUIESCENCE")])
mitf_geneSig <- lapply(mitf_geneSig, function(x) x[x != ""])

# now take the hallmark pathway gene signatures
hallmark_pathway_geneSig <- read.csv("/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/Project2_stable_disease_project/GeoMX_analysis/hallmark_pathway.csv")[-1]
hallmark_pathway_geneSigList <- lapply(as.list(hallmark_pathway_geneSig), function(x) x[x != ""])
#hallmark_pathway_geneSigList[0:1]

colnames(all_tumorSig) <- make.names(colnames(all_tumorSig))

all_tumorSig_list <- lapply(as.list(all_tumorSig), function(x) x[!is.na(x)])
```

```{r, all tumor signatures}
all_states <- c("Tirosh.MITF", "Tirosh.AXL", 
                "Tsoi.UNDIFF", "Tsoi.NEURAL.CREST.like", "Tsoi.TRANSITORY", "Tsoi.MELANOCYTIC", 
                "Wouters.INTERMEDIATE", "Wouters.MES.like", "Wouters.MELANOCYTIC", 
                "Baron.STRESS", "Baron.NC.like", "Baron.MELANOCYTIC",
                "Rambow.NCSC", "Rambow.INVASIVE", "Rambow.SMC", "Rambow.HYPERdiff",
                "Karras.melanocytic..Oxphos.", "Karras.Neural.Crest.like", "Karras.RNA.processing", "Karras.Antigen.presentation", "Karras.Stem.like", "Karras.Stess.like..hypoxia.", "Karras.mesenchymal.like")


selected_tumor_states <- c("Tirosh.MITF", "Tirosh.AXL", 
                "Tsoi.NEURAL.CREST.like", "Tsoi.MELANOCYTIC", 
                "Wouters.MES.like", "Wouters.MELANOCYTIC", 
                "Baron.NC.like", "Baron.MELANOCYTIC",
                "Rambow.NCSC", "Rambow.HYPERdiff",
                "Karras.melanocytic..Oxphos.", "Karras.Neural.Crest.like")

tumor_states <- all_tumorSig_list[all_states]
selected_states <- all_tumorSig_list[selected_tumor_states]

summary(tumor_states)

names(selected_states) <- c("Tirosh.MITF(100)", "Tirosh.AXL(100)", 
                "Tsoi.NEURAL.CREST.like(66)", "Tsoi.MELANOCYTIC(62)", 
                "Wouters.MES.like(2115)", "Wouters.MELANOCYTIC(1574)", 
                "Baron.NC.like(212)", "Baron.MELANOCYTIC(176)",
                "Rambow.NCSC(42)", "Rambow.HYPERdiff(15)",
                "Karras.melanocytic.Oxphos(17)", "Karras.Neural.Crest.like(26)")

names(tumor_states) <- c("Tirosh.MITF(100)", "Tirosh.AXL(100)", "Tsoi.UNDIFF(118)",
                         "Tsoi.NEURAL.CREST.like(66)", "Tsoi.TRANSITORY(29)", "Tsoi.MELANOCYTIC(62)",
                         "Wouters.INTERMEDIATE(394)", "Wouters.MES.like(2115)", "Wouters.MELANOCYTIC(1574)", 
                         "Baron.STRESS(61)", "Baron.NC.like(212)", "Baron.MELANOCYTIC(176)",
                         "Rambow.NCSC(42)", "Rambow.SMC(14)", "Rambow.HYPERdiff(15)", 
                         "Karras.melanocytic.Oxphos(17)", "Karras.Neural.Crest.like(26)", "Karras.RNA.processing(32)", "Karras.Antigen.presentation(42)",
                         "Karras.Stem.like(38)", "Karras.Stess.like.hypoxia(20)", "Karras.mesenchymal.like(32)")
```
actual analysis for entrop plotting
```{r, fig.height = 6, fig.width = 8}
merged_entropy <- read.csv("/Users/yshi797/Desktop/PCA/merged_entropy.csv", row.names = 1)
merged_entropy

custom_colors <- c(
  "N" = "skyblue",  # orange
  "MIS" = "purple2",    # sky blue
  "P" = "darkgreen",      # teal
  "RGP" = "orange", # yellow
  "VGP" = "brown3"
  #"DN" = "grey"# blue
)

text_size = 12

merged_entropy_plot <- ggplot(merged_entropy, aes(PC1, PC2, fill = histopathology)) +
  geom_point(shape = 21, size = 2, alpha = 0.7, color = "black") +
  labs(
    #x = paste0("PC1: ", percentVar_non[1], "% variance"),
   #    y = paste0("PC2: ", percentVar_non[2], "% variance"),
       title = "PCA Plot colored by histopathology and shaped by batch") +
  scale_fill_manual(values = custom_colors) +  # Use the defined custom colors and fallback color
  theme_classic() +
  theme(legend.position = "bottom",
        text = element_text(size = text_size),
        axis.text = element_text(size = text_size),
        axis.line = element_line(size = 0.5, color = "black"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank())

print(merged_entropy_plot)
```



# now define the comparison groups based on PC1
```{r}
merged_entropy$comparison <- "Others"
merged_entropy[merged_entropy$PC1 > 0, "comparison"] <- "right_PC1"
merged_entropy[merged_entropy$PC1 <= 0, "comparison"] <- "left_PC1"

table(merged_entropy$comparison)


# Create the boxplot
entropy_plot <- ggplot(merged_entropy, aes(x = comparison, y = H_spatial#, fill = histopathology
                                           )) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +  # Boxplot without default outliers
  geom_jitter(width = 0.2, size = 2, color = "grey50", alpha = 0.7) +  # Jittered points
  #scale_x_discrete(limits = x_order) +  # Set custom order for x-axis
  scale_fill_manual(values = custom_colors) +  # Set custom fill colors
  theme_minimal() +
  labs(
    title = "Comparison of Spatial Entropy by Breslow Bucket",
    x = "Breslow Bucket",
    y = "Spatial Entropy"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.position = "none"
  )

print(entropy_plot)
```


```{r}
merged_entropy_sub <- merged_entropy[merged_entropy$histopathology %in% c("VGP", "RGP"), ]

merged_entropy_sub$comparison <- "Others"
merged_entropy_sub[merged_entropy_sub$PC1 > 0, "comparison"] <- "right_PC1"
merged_entropy_sub[merged_entropy_sub$PC1 <= 0, "comparison"] <- "left_PC1"

table(merged_entropy_sub$comparison)


# Create the boxplot
entropy_plot <- ggplot(merged_entropy_sub, aes(x = comparison, y = H_spatial#, fill = histopathology
                                           )) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +  # Boxplot without default outliers
  geom_jitter(width = 0.2, size = 2, color = "grey50", alpha = 0.7) +  # Jittered points
  #scale_x_discrete(limits = x_order) +  # Set custom order for x-axis
  scale_fill_manual(values = custom_colors) +  # Set custom fill colors
  theme_minimal() +
  labs(
    title = "Comparison of Spatial Entropy by Breslow Bucket",
    x = "Breslow Bucket",
    y = "Spatial Entropy"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.position = "none"
  )

print(entropy_plot)
```



# Access the data
```{r, read the data}
PCAI_merged_pheno <- read.csv("/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/processed_data/PCAI_merged_pheno_05222024.csv", row.names = 1)
PCAI_merged_raw <- read.csv("/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/processed_data/PCAI_merged_raw_05192024.csv", row.names = 1)
PCAI_merged_vst <- read.csv("/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/processed_data/PCAI_merged_norm_batchcorrected_05192024.csv", row.names = 1)
```

```{r, clear the rownames for matching}
# Define a function to replace only the first three dots with hyphens
replace_first_three_dots <- function(name) {
  # Split the name by dots
  parts <- unlist(strsplit(name, "\\."))
  new_name <- paste(paste(parts[1:4], collapse = "-"), paste(parts[5:length(parts)], collapse = "."), sep = ".")
  return(new_name)
  }

colnames(PCAI_merged_raw) <- sapply(colnames(PCAI_merged_raw), replace_first_three_dots)
colnames(PCAI_merged_vst) <- sapply(colnames(PCAI_merged_vst), replace_first_three_dots)
```

# check the data matching 
```{r}
head(colnames(PCAI_merged_raw))
head(colnames(PCAI_merged_vst))
```
# to run SOM use log transformed data. 
# use tumor ROI only sample
# identify t

```{r}
print(table(PCAI_merged_pheno$histopathology))
print(table(PCAI_merged_pheno$composition_TV))
selected_hist <- c("N", "P", "MIS", "VGP", "RGP")
selected_celltype <- c("tumor", "melanocyte")
```
```{r, sample selection}
batch2_tumor_samples <- rownames(PCAI_merged_pheno[(PCAI_merged_pheno$batch_str == "Batch_2") &
                                                     (PCAI_merged_pheno$histopathology %in% selected_hist) &
                                                     (PCAI_merged_pheno$composition_TV %in% selected_celltype), ])
#print(batch2_tumor_samples)
print(length(batch2_tumor_samples))
```

# check on batch 1 data
```{r}
batch1_tumor_samples <- rownames(PCAI_merged_pheno[(PCAI_merged_pheno$batch_str == "Batch_1") &
                                                     (PCAI_merged_pheno$histopathology %in% selected_hist) &
                                                     (PCAI_merged_pheno$composition_TV %in% selected_celltype), ])
print(length(batch1_tumor_samples))

tumor_pheno_b1 <- PCAI_merged_pheno[batch1_tumor_samples, ]
tumor_raw_count_b1 <- PCAI_merged_raw[, batch1_tumor_samples]

table(tumor_pheno_b1$histopathology)

#tumor_raw_count_b1
# now create dds project
b1_dds <- DESeqDataSetFromMatrix(countData = round(tumor_raw_count_b1),
                                 colData = tumor_pheno_b1,
                                 design = ~ histopathology)

b1_dds <- DESeq(b1_dds)
# extract the normalized counts
b1_dds <- estimateSizeFactors(b1_dds)
b1_norm <- counts(b1_dds, normalize = T)
b1_log <- log2(b1_norm + 1)
#b2_log_matrix <- as.matrix(b2_log)

table(tumor_pheno_b1$composition_TV)
```
```{r, SOM input prep for batch 1}
# assign input
b1_exprs_data <- b1_log

b1_group.info <- data.frame(
  group.labels = tumor_pheno_b1$histopathology,
  row.names = row.names(tumor_pheno_b1)
)

b1_group.info$group.colors = "unknown"

b1_group.info[b1_group.info$group.labels == "P", "group.colors"] <- "forestgreen"
b1_group.info[b1_group.info$group.labels == "MIS", "group.colors"] <- "pink"
b1_group.info[b1_group.info$group.labels == "RGP", "group.colors"] <- "orange"
b1_group.info[b1_group.info$group.labels == "VGP", "group.colors"] <- "tomato"

print(table(b1_group.info$group.labels))
print(table(b1_group.info$group.colors))

b1_pheno_data_color <- AnnotatedDataFrame(data = b1_group.info)

b1_exprs_set <- ExpressionSet(assayData = b1_exprs_data,
                              phenoData = b1_pheno_data_color)

b1_exprs_set
```

```{r, create SOM envrionment for batch1 tumor using 30}
b1_env <- opossom.new(list(dataset.name = "PCAI_batch1_tumor_30x30_01212025", dim.1stLvlSom = 30))
b1_env$indata <- b1_exprs_set
b1_env$group.labels <- b1_group.info$group.labels
b1_env$group.colors <- b1_group.info$group.colors

b1_env
```

```{r, run SOM for batch1 tumor}
#opossom.run(b1_env)
```


```{r}
b1_env_40 <- opossom.new(list(dataset.name = "PCAI_batch1_tumor_40x40_01212025", dim.1stLvlSom = 40))
b1_env_40$indata <- b1_exprs_set
b1_env_40$group.labels <- b1_group.info$group.labels
b1_env_40$group.colors <- b1_group.info$group.colors

#opossom.run(b1_env_40)
```


# subset the data and create the DEG project (for generating nomalized log-transformed counts)
```{r, dds creation}
tumor_pheno_b2 <-  PCAI_merged_pheno[batch2_tumor_samples, ]
tumor_raw_count_b2 <- PCAI_merged_raw[, batch2_tumor_samples]

# now create dds project
b2_dds <- DESeqDataSetFromMatrix(countData = round(tumor_raw_count_b2),
                                 colData = tumor_pheno_b2,
                                 design = ~ histopathology)
# extract the normalized counts
b2_dds <- DESeq(b2_dds)
b2_dds <- estimateSizeFactors(b2_dds)
b2_norm <- counts(b2_dds, normalize = T)
b2_log <- log2(b2_norm + 1)
#b2_log_matrix <- as.matrix(b2_log)
```



# now rerun SOM
```{r, SOM input prep 2}
# assign input
b2_exprs_data <- b2_log

b2_group.info <- data.frame(
  group.labels = tumor_pheno_b2$histopathology,
  row.names = row.names(tumor_pheno_b2)
)

# assign color to each label
b2_group.info$group.colors = "unknown"

b2_group.info[b2_group.info$group.labels == "N", "group.colors"] <- "steelblue"
b2_group.info[b2_group.info$group.labels == "P", "group.colors"] <- "forestgreen"
b2_group.info[b2_group.info$group.labels == "MIS", "group.colors"] <- "pink"
b2_group.info[b2_group.info$group.labels == "RGP", "group.colors"] <- "orange"
b2_group.info[b2_group.info$group.labels == "VGP", "group.colors"] <- "tomato"

print(table(b2_group.info$group.labels))
print(table(b2_group.info$group.colors))

b2_pheno_data_color <- AnnotatedDataFrame(data = b2_group.info)
#pheno_data_color
b2_exprs_set <- ExpressionSet(assayData = b2_exprs_data,
                              phenoData = b2_pheno_data_color)

b2_exprs_set
```

```{r, create SOM envrionment 2}
b2_env_40 <- opossom.new(list(dataset.name = "PCAI_batch2_tumor_40x40_01212024", dim.1stLvlSom = 40))
b2_env_40$indata <- b2_exprs_set
b2_env_40$group.labels <- b2_group.info$group.labels
b2_env_40$group.colors <- b2_group.info$group.colors
```

```{r, run SOM 2}
#opossom.run(b2_env_40)
```
# now extract 2 separate branch
```{r}
# reference code is from sample_similarity_analysis_corr.r  
# https://github.com/hloefflerwirth/oposSOM/blob/master/R/sample_similarity_analysis_cor.r

mixColors<-function(c1, c2,alpha=122 )
  {
    do.call( rgb,
             as.list( c( colMeans( sapply(c(2,4,6),function(i)
               c( strtoi(substr(c1,i,i+1),16), strtoi(substr(c2,i,i+1),16) )
               
             ) ), alpha ) /255 )
    )
}

groupwise.group.colors.stability.1 <- apply((col2rgb(b2_env$groupwise.group.colors) + 0.7 * (255 - col2rgb(b2_env$groupwise.group.colors))) / 255,2, function(x) rgb(x[1],x[2],x[3]))

groupwise.group.colors.stability.2 <- apply((col2rgb(b2_env$groupwise.group.colors) + 0.9 *
          (255 - col2rgb(b2_env$groupwise.group.colors))) / 255,
       2, function(x) rgb(x[1],x[2],x[3]))

stability.colors <- b2_env$group.colors
  stability.colors[which(b2_env$group.silhouette.coef < 0.5)] <- groupwise.group.colors.stability.1[b2_env$group.labels[names(which(b2_env$group.silhouette.coef < 0.5))]]
  stability.colors[which(b2_env$group.silhouette.coef < 0)] <- groupwise.group.colors.stability.2[b2_env$group.labels[names(which(b2_env$group.silhouette.coef < 0))]]

  
pcm.module <- cor(b2_env[[paste("spot.list.", b2_env$preferences$standard.spot.modules,sep="")]]$spotdata )
pcm.metadata <- cor(b2_env$metadata)
```

```{r, recapitulate the spanning tree, fig.height= 6, fig.width= 6}
# Generate the tree only for metadata
adj.matrix <- pcm.metadata
adj.matrix[which(is.na(adj.matrix))] <- 0
n <- "metagene data"

# Create the graph from the adjacency matrix
g <- graph.adjacency(-adj.matrix, weighted=TRUE, mode="undirected")
stg <- minimum.spanning.tree(g)
E(stg)$weight <- 1

# Layout for the spanning tree
layout <- layout_with_kk(stg)

# Extract x and y locations
xy_locations <- data.frame(x = layout[, 1], y = layout[, 2])

# Plot the spanning tree
par(mar=c(1,1,1,1))

E(stg)$color <- apply(get.edgelist(stg), 1, function(x) mixColors(b2_env$group.colors[x[1]], b2_env$group.colors[x[2]]))

plot(stg, layout=layout, vertex.size=5, vertex.label=rep("", ncol(adj.matrix)),
     vertex.color=b2_env$group.colors, main=paste("Correlation Spanning Tree on", n))

legend("bottomright", as.character(unique(b2_env$group.labels)), cex=0.5, text.col=b2_env$groupwise.group.colors, bg="white")

box()
```
```{r}
# Extract vertex names (sample labels)
vertex_names <- V(stg)$name

# Create a data frame with sample labels and their corresponding x and y coordinates
xy_locations <- data.frame(
  sample_label = vertex_names,
  x = layout[, 1],
  y = layout[, 2]
)

# View the matched data frame
print(xy_locations)
```
```{r}
# Create a data frame with sample labels and their corresponding x and y coordinates
xy_locations <- data.frame(
  sample_label = vertex_names,
  x = layout[, 1],
  y = layout[, 2]
)

# View the matched data frame
print(xy_locations)

```
```{r, fig.height = 6, fig.width= 6}
# Assuming xy_locations is your data frame
ggplot(xy_locations, aes(x = x, y = y, label = sample_label)) +
  geom_point(color = "blue", size = 3) +  # Points for each sample
  #geom_text(vjust = -0.5, hjust = 0.5, size = 3) +  # Labels for each point
  theme_minimal() +
  labs(title = "Correlation Spanning Tree - Sample Locations",
       x = "X Coordinate",
       y = "Y Coordinate") +
  scale_x_continuous(breaks = seq(floor(min(xy_locations$x)), ceiling(max(xy_locations$x)), by = 0.5)) +
  scale_y_continuous(breaks = seq(floor(min(xy_locations$y)), ceiling(max(xy_locations$y)), by = 0.5)) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )
```



# combine the data with pheno data and assign new annotation
```{r, assign_invasive branchs}
branch2 <- xy_locations[(xy_locations$x >= -1.75) &
                                  (xy_locations$x <= 0) &
                                   (xy_locations$y >= 0) &
                                   (xy_locations$y <= 3), ]$sample_label
branch1 <- xy_locations[(xy_locations$x <= -0.5) & 
                                   (xy_locations$y >= -1.5), ]$sample_label

branch3 <- xy_locations[(xy_locations$x > 0) & 
                                   (xy_locations$y >= -0.5), ]$sample_label

# clean up branch1
overlap <- intersect(branch1, branch2)
overlap
#

xy_locations$branch <- "others"
xy_locations[xy_locations$sample_label %in% branch1, "branch"] <- "branch1"
xy_locations[xy_locations$sample_label %in% branch2, "branch"] <- "branch2"
xy_locations[xy_locations$sample_label %in% branch3, "branch"] <- "branch3"

table(xy_locations$branch)
```

```{r, fig.height= 6, fig.width= 8}
# Assuming xy_locations is your data frame
ggplot(xy_locations, aes(x = x, y = y, color = branch)) +
  geom_point(size = 4) +  # Points for each sample
  #geom_text(vjust = -0.5, hjust = 0.5, size = 3) +  # Labels for each point
  theme_minimal() +
  labs(title = "Correlation Spanning Tree - Sample Locations",
       x = "X Coordinate",
       y = "Y Coordinate") +
  scale_x_continuous(breaks = seq(floor(min(xy_locations$x)), ceiling(max(xy_locations$x)), by = 0.5)) +
  scale_y_continuous(breaks = seq(floor(min(xy_locations$y)), ceiling(max(xy_locations$y)), by = 0.5)) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )
```
# at the end rerun the inferCNV for batch2 data only
```{r, combine the data with pheno}
tumor_pheno_b2$sample_label <- rownames(tumor_pheno_b2)
# b2_pheno contains information about different branches
b2_pheno <- merge(tumor_pheno_b2, xy_locations, by = "sample_label")
table(b2_pheno$branch)

# rename x and y colnames
b2_pheno <- b2_pheno %>% 
  rename(spanning_tree_x = x, spanning_tree_y = y)

rownames(b2_pheno) <- b2_pheno$sample_label
```
# now recreate dds object with comparison within branchs
```{r, dds creation with branches}
# now create dds project
b2_branch_dds <- DESeqDataSetFromMatrix(countData = round(tumor_raw_count_b2),
                                 colData = b2_pheno,
                                 design = ~ branch)
# extract the normalized counts
b2_branch_dds <- estimateSizeFactors(b2_branch_dds)
b2_branch_norm <- counts(b2_branch_dds, normalize = T)
b2_branch_log <- log2(b2_branch_norm + 1)
```


# now rerun SOM
```{r, SOM input prep}
# assign input
exprs_data <- b2_branch_log

group.info <- data.frame(
  group.labels = b2_pheno$branch,
  row.names = row.names(b2_pheno)
)

# assign color to each label
group.info$group.colors = "unknown"

group.info[group.info$group.labels == "others", "group.colors"] <- "steelblue"
group.info[group.info$group.labels == "invasive_branch1", "group.colors"] <- "pink"
group.info[group.info$group.labels == "invasive_branch2", "group.colors"] <- "orange"
group.info[group.info$group.labels == "invasive_branch3", "group.colors"] <- "tomato"

print(table(group.info$group.labels))
print(table(group.info$group.colors))

pheno_data_color <- AnnotatedDataFrame(data = group.info)
#pheno_data_color
b2_exprs_set <- ExpressionSet(assayData = exprs_data,
                              phenoData = pheno_data_color)

b2_exprs_set

```

```{r, create SOM envrionment}
b2_branch_env <- opossom.new(list(dataset.name = "PCAI_batch2_tumor_branches_30x30_09082024", dim.1stLvlSom = 30))
b2_branch_env$indata <- b2_exprs_set
b2_branch_env$group.labels <- group.info$group.labels
b2_branch_env$group.colors <- group.info$group.colors
```

```{r, run SOM}
library(oposSOM)
#opossom.run(b2_branch_env)
```

```{r assign annotation to different branches}
# add the original annotation 
ori_merged_pheno <- read.csv("/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/GeoMX_data/PCAI_merged_pheno_05222024.csv", row.names = 1)

ori_merged_tumor_b2 <- ori_merged_pheno[(ori_merged_pheno$batch_str == "Batch_2") & 
                                          (ori_merged_pheno$composition_TV %in% selected_celltype) &
                                          (ori_merged_pheno$histopathology %in% selected_hist), ]

ori_merged_tumor_b2$sample_label <- rownames(ori_merged_tumor_b2)


```

```{r, output the data with branch annotation}
b2_pheno_merged <- merge(ori_merged_tumor_b2, b2_pheno[, c("sample_label", "spanning_tree_x", "spanning_tree_y", "branch")], by = "sample_label")
b2_pheno_merged$slide_rois <- paste0(b2_pheno_merged$slide.name, "_", b2_pheno_merged$roi_label)
b2_pheno_merged

#write.csv(b2_pheno_merged, "/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/script/PCAI_batch2_tumor_SOM_branches_pheno_09082024.csv", row.names = FALSE)

```
# match the data with clusters
```{r input cluster annotation}
cluster_anno <- read.csv("/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/processed_data/invasive_tumor_GeoMX_cluster.csv", row.names = 1)[, c("name", "cluster", "batch")]
cluster_anno <- cluster_anno %>% 
  rename(sample_label = name) 

cluster_anno_b2 <- cluster_anno[cluster_anno["batch"] == "Batch_2", ]
```

# be carefull with the data merge
```{r}
in_cluster1 <- rownames(cluster_anno_b2[cluster_anno_b2$cluster == "1", ])
in_cluster2 <- rownames(cluster_anno_b2[cluster_anno_b2$cluster == "2", ])
in_cluster3 <- rownames(cluster_anno_b2[cluster_anno_b2$cluster == "3", ])

b2_pheno_merged$cluster = "others"
b2_pheno_merged[b2_pheno_merged$sample_label %in% in_cluster1, "cluster"] = "cluster1"
b2_pheno_merged[b2_pheno_merged$sample_label %in% in_cluster2, "cluster"] = "cluster2"
b2_pheno_merged[b2_pheno_merged$sample_label %in% in_cluster3, "cluster"] = "cluster3"

table(b2_pheno_merged$cluster)

#cluster_merge <- merge(b2_pheno_merged, cluster_anno_b2, by = "sample_label", all.x = T)
#cluster_merge$cluster <- as.character(cluster_merge$cluster)
```

```{r, fig.height= 6, fig.width= 8}
# Assuming xy_locations is your data frame
ggplot(b2_pheno_merged, aes(x = spanning_tree_x, y = spanning_tree_y, color = cluster)) +
  geom_point(size = 4) +  # Points for each sample
  #geom_text(vjust = -0.5, hjust = 0.5, size = 3) +  # Labels for each point
  theme_minimal() +
  labs(title = "Correlation Spanning Tree - Sample Locations",
       x = "X Coordinate",
       y = "Y Coordinate") +
  scale_x_continuous(breaks = seq(floor(min(xy_locations$x)), ceiling(max(xy_locations$x)), by = 0.5)) +
  scale_y_continuous(breaks = seq(floor(min(xy_locations$y)), ceiling(max(xy_locations$y)), by = 0.5)) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )
```

# the analysis starts from here


# starting from here, read the annotation data
```{r}
b2_pheno_merged <- read.csv("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/PCAII_project_manuscript_preparation_2025/processed_data/PCAI_Set1A_tumor_SOM_branches_pheno_09082024.csv")

rownames(b2_pheno_merged) <- b2_pheno_merged$sample_label
b2_pheno_merged
```



```{r, fig.height= 6, fig.width= 8}
# Define the specific values in 'slide_rois' to label
label_values <- c("LSP13943_11", "LSP13943_14")  # Replace with the actual values

ggplot(b2_pheno_merged, aes(x = spanning_tree_x, y = spanning_tree_y, color = branch)) +
  geom_point(size = 4) +  # Points for each sample
  geom_text(
    data = b2_pheno_merged %>% dplyr::filter(slide_rois %in% label_values),  # Subset only rows to label
    aes(label = slide_rois),  # Use 'slide_rois' column for labels
    vjust = -0.5, hjust = 0.5, size = 3, color = "black"  # Adjust position and size
  ) +
  theme_minimal() +
  labs(
    title = "Correlation Spanning Tree - Sample Locations",
    x = "X Coordinate",
    y = "Y Coordinate"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )
```

# now assign invasive branches which are RGP and VGP regions
```{r invasive branches preparation}
invasive_regions <- c("RGP", "VGP")
table(b2_pheno_merged$histopathology)

invasive_branch1 <- b2_pheno_merged[(b2_pheno_merged$branch == "branch1") & (b2_pheno_merged$histopathology %in% invasive_regions), ]$sample_label
invasive_branch2 <- b2_pheno_merged[(b2_pheno_merged$branch == "branch3") & (b2_pheno_merged$histopathology %in% invasive_regions), ]$sample_label

# check the length of the samples
print(length(invasive_branch1))
print(length(invasive_branch2))

invasive_pheno <- b2_pheno_merged[rownames(b2_pheno_merged) %in% c(invasive_branch1, invasive_branch2), ]
invasive_pheno

# now subset the rawcount
invasive_raw <- tumor_raw_count_b2[, c(invasive_branch1, invasive_branch2)]
invasive_raw

table(invasive_pheno$branch)

invasive_pheno <- invasive_pheno[match(colnames(invasive_raw), rownames(invasive_pheno)), ]

all(colnames(invasive_raw) == rownames(invasive_pheno))  # Should return TRUE
```


```{r}
#sample_list <- colnames(invasive_raw)
branch1_samples <- rownames(invasive_pheno[invasive_pheno$branch == "branch1", ])
branch3_samples <- rownames(invasive_pheno[invasive_pheno$branch == "branch3", ])

cibersort_tpm <- invasive_raw %>% 
  rownames_to_column(var = "GeneSymbol") 
  #%>% 
  #rename_with(~ ifelse(. %in% branch1_samples, "branch1",
   #                    ifelse(. %in% branch3_samples, "branch3", .)),
    #          -GeneSymbol)

# Get the current column names excluding 'GeneSymbol'
current_sample_names <- colnames(cibersort_tpm)[-1]

# Replace sample names with 'branch1' or 'branch3' based on their group
# Initialize new column names by matching samples to branch1 and branch3
new_column_names <- sapply(current_sample_names, function(sample) {
  if (sample %in% branch1_samples) {
    return("branch1")
  } else if (sample %in% branch3_samples) {
    return("branch3")
  } else {
    return(sample)  # Keep the original name if it doesn't belong to branch1 or branch3
  }
})

# Assign the new names back to the data (preserving repeated names)
colnames(cibersort_tpm)[-1] <- new_column_names

# Check the result
print(cibersort_tpm)
  
#write.table(cibersort_tpm, file = "branch_cibersort_signature_matrix_09302024.txt", sep = "\t", row.names =
 #             FALSE, quote = FALSE)

```




```{r}
# now creating DEG object 
invasive_dds <-  DESeqDataSetFromMatrix(countData = round(invasive_raw),
                                 colData = invasive_pheno,
                                 design = ~ branch)

invasive_dds <- DESeq(invasive_dds)
vsd <- vst(invasive_dds, blind = FALSE)
# comparison
res <- results(invasive_dds, contrast = c("branch", "branch1", "branch3"))
res <- res[order(res$padj), ]
significant_genes <- rownames(res[which(res$padj <= 0.05), ])

```


# extract signatures
```{r}
# Only consider significant genes and then sort by log2FoldChange
significant_res <- res[significant_genes, ]

# Get top 100 upregulated genes for branch1 (log2FoldChange > 0)
branch1_signature <- rownames(significant_res[significant_res$log2FoldChange > 0.7, ][1:100, ])  

# Get top 100 upregulated genes for branch3 (log2FoldChange < 0)
branch3_signature <- rownames(significant_res[significant_res$log2FoldChange < -0.6, ][1:100, ])
# Map gene symbols to Entrez IDs
# Create a list of lists with names
gene_signatures <- list(
  branch1_signature = branch1_signature,
  branch3_signature = branch3_signature
)

# Assign meaningful names to each list element like the example
names(gene_signatures) <- c("BRANCH1_SIGNATURE", "BRANCH3_SIGNATURE")

branch1_signature

library(EnsDb.Hsapiens.v79)
```



```{r}
# now extract the normalized count data
invasive_dds <- estimateSizeFactors(invasive_dds)
invasive_norm <- counts(invasive_dds, normalize = T)
invasive_log <-  as.data.frame(log2(invasive_norm + 1))
invasive_log_matrix <- as.matrix(invasive_log)

```



```{r}
invasive_log_matrix 
source("test_script.R")
source("score_gene_expression.R")
# read the markers
#human_DGM_markers <- readRDS("Human_DGM_Markers.rds")
score_info <- compute_bulk_normalized_gene_set_scores(invasive_log_matrix, dgms)
score_info

```

```{r}
dgm_score_mat <- score_info$fg - score_info$bg
dgm_score_mat
```
# generate boxplot comparison for each of them 

```{r}
dgm_score_df <- as.data.frame(dgm_score_mat)
dgm_score_df
```


```{r}
ATF_expr <- as.data.frame(t(invasive_log))

ATF_expr

comparison <- merge(invasive_pheno, ATF_expr, by = "row.names")

rownames(comparison) <- comparison$Row.names

comparison$Row.names <- NULL
```

# generate boxplot
```{r}
# Create the boxplot and stripplot for each column
p <- ggplot(comparison, aes(x = branch, y = ATF3, fill = branch)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +    # Boxplot without outliers
  geom_jitter(shape = 16, position = position_jitter(0.2), size = 3) +  # Stripplot
  scale_fill_manual(values = c("branch1" = "pink", "branch3" = "skyblue")) +  # Customcolors for branches
  stat_compare_means(method = "wilcox.test", label = "p.format", label.y = max(comparison$ATF3, na.rm = TRUE) * 1.05) +  # Add p-value
  theme_minimal() +
  labs(title = paste("Comparison of", column, "\n   between branch1 and branch3"),
       y = column, x = "Branch") +
  theme(legend.position = "none")  # Hide legend for fill

  # Save the plot
print(p)
```


```{r}
# check MEL27
comparison_mel27 <- comparison[comparison$case == "MEL27", ]
comparison_mel27

table(comparison_mel27$slide_rois)

#cluster1 <- c("LSP13914_6", "LSP13914_7", "LSP13914_8", "LSP13914_9", "LSP13914_12")
cluster3 <- c("LSP13914_6", "LSP13914_7", "LSP13914_8", "LSP13914_9", "LSP13914_11", "LSP13914_12", "LSP13914_5")


cluster1 <- c("LSP13918_43", "LSP13918_45", "LSP13914_14", "LSP13918_29", "LSP13918_6", 
              "LSP13918_7", "LSP13918_31")


comparison_mel27$comparison <- "cluster2"
comparison_mel27[comparison_mel27$slide_rois %in% cluster3, "comparison"] <- "cluster3"
comparison_mel27[comparison_mel27$slide_rois %in% cluster1, "comparison"] <- "cluster1"


table(comparison_mel27$comparison)

```

```{r, fig.height= 6, fig.width= 4}
# Create the boxplot and stripplot for each column
p_ATF3 <- ggplot(comparison_mel27, aes(x = comparison, y = ATF3, fill = comparison)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +    # Boxplot without outliers
  geom_jitter(shape = 16, position = position_jitter(0.2), size = 3) +  # Stripplot
  scale_fill_manual(values = c("cluster3" = "pink", "cluster1" = "skyblue")) +  # Customcolors for branches
  stat_compare_means(method = "wilcox.test", label = "p.format", label.y= max(comparison_mel27$ATF3, na.rm = TRUE) * 1.05) +  # Add p-value
  theme_minimal() +
  labs(title = paste("Comparison of ATF3\n   between cluster1 and cluster3"),
       y = "ATF3(log-transformed)", x = "Branch") +
  theme(legend.position = "none")  # Hide legend for fill


# Create the boxplot and stripplot for each column
p_NGFR <- ggplot(comparison_mel27, aes(x = comparison, y = NGFR, fill = comparison)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +    # Boxplot without outliers
  geom_jitter(shape = 16, position = position_jitter(0.2), size = 3) +  # Stripplot
  scale_fill_manual(values = c("cluster3" = "pink", "cluster1" = "skyblue")) +  # Customcolors for branches
  stat_compare_means(method = "wilcox.test", label = "p.format", label.y= max(comparison_mel27$NGFR, na.rm = TRUE) * 1.05) +  # Add p-value
  theme_minimal() +
  labs(title = paste("Comparison of NGFR\n between cluster1 and cluster3"),
       y = "NGFR(log-transformed)", x = "Branch") +
  theme(legend.position = "none")  # Hide legend for fill




# Save the plot
print(p_ATF3)
print(p_NGFR)


```

# more geneset analysis

```{r}
# read Ponzniak signatures
ponzniak_geneset <- read_csv("Ponzniak_signatures_2024.csv")
table(ponzniak_geneset$cluster)
```
```{r}
ponzniak_list <- split(ponzniak_geneset$gene, ponzniak_geneset$cluster)

ponzniak_list <- ponzniak_list[!names(ponzniak_list) %in% c("Patient_specific_A", "Patient_specific_B")]

# now perform gsva
```

```{r}
ponzniak_gsva <- gsva(ssgseaParam(invasive_log_matrix, ponzniak_list), verbose = F)
ponzniak_gsvaDF <- as.data.frame(t(ponzniak_gsva))
ponzniak_gsvaDF
```

```{r, fig.height = 8, fig.width = 12}
breaks_list <- seq(-2, 2, by = 0.05)
print(length(breaks_list))
color_scale <- colorRampPalette(c("navy", "white", "red"))(length(breaks_list))

cl_cb <- function(hcl, mat){
    # Recalculate manhattan distances for reorder method
    dists <- dist(ponzniak_gsvaDF, method = "manhattan")

    # Perform reordering according to OLO method
    hclust_olo <- reorder(hcl, dists)
    return(hclust_olo)
}

# Start SVG device
#pdf("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/ponzniak_signature_analysis_heatmap.pdf", width = 12, height = 8)

pheatmap(t(ponzniak_gsvaDF),
         scale = "row",
         border_color = "grey",
         show_colnames = F,
         cellheight = 10,
         cellwidth = 3,
         treeheight_row = 15, 
         treeheight_col = 25,
         color = color_scale, 
         breaks = breaks_list, 
         cluster_cols = TRUE, 
         cluster_rows = FALSE,
         clustering_callback = cl_cb,
         annotation_col= invasive_pheno[, c("histopathology", "Immune_TV", "branch")],
         annotation_colors = list(branch = c("branch1" = "tomato",
                                            "branch3" = "steelblue")),
          #                        case = mel_colors),
         fontsize = 12,         # General font size
         fontsize_row = 10,     # Font size for row names
         fontsize_col = 10,   
         legend = T,
         annotation_legend = TRUE)

#dev.off()
```
now try different method
```{r}
ponzniak_info <- compute_bulk_normalized_gene_set_scores(invasive_log_matrix, ponzniak_list)
ponzniak_info
```
```{r, fig.height= 10, fig.width= 12}
ponzniak_mat <- ponzniak_info$fg - ponzniak_info$bg
ponzniakDF <- as.data.frame(ponzniak_mat)
```

```{r, fig.height = 8, fig.width = 12}
breaks_list <- seq(-2, 2, by = 0.05)
print(length(breaks_list))
color_scale <- colorRampPalette(c("navy", "white", "red"))(length(breaks_list))

cl_cb <- function(hcl, mat){
    # Recalculate manhattan distances for reorder method
    dists <- dist(t(ponzniakDF), method = "manhattan")

    # Perform reordering according to OLO method
    hclust_olo <- reorder(hcl, dists)
    return(hclust_olo)
}

# Start SVG device
pdf("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/ponizniak_signature_collector_function_analysis_heatmap.pdf", width = 12, height = 8)

pheatmap(ponzniakDF,
         scale = "row",
         border_color = "grey",
         show_colnames = F,
         cellheight = 10,
         cellwidth = 3,
         treeheight_row = 15, 
         treeheight_col = 25,
         color = color_scale, 
         breaks = breaks_list, 
         cluster_cols = TRUE, 
         cluster_rows = F,
         clustering_callback = cl_cb,
         annotation_col= invasive_pheno[, c("histopathology", "Immune_TV", "branch")],
         annotation_colors = list(branch = c("branch1" = "tomato",
                                            "branch3" = "steelblue")),
          #                        case = mel_colors),
         fontsize = 12,         # General font size
         fontsize_row = 10,     # Font size for row names
         fontsize_col = 10,   
         legend = T,
         annotation_legend = TRUE)

dev.off()
```

```{r}
ponzniak_gsvaDF_merged <- merge(ponzniak_gsvaDF, invasive_pheno, by = "row.names")
ponzniakDF_merged <- merge(as.data.frame(t(ponzniakDF)), invasive_pheno, by = "row.names")
```


```{r}
ponzniak_gsvaDF_merged
```



```{r}
# Create the boxplot and stripplot for each column
p <- ggplot(ponzniak_gsvaDF_merged, aes(x = branch, y = Melanocytic, fill = branch)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +    # Boxplot without outliers
  geom_jitter(shape = 16, position = position_jitter(0.2), size = 3) +  # Stripplot
  scale_fill_manual(values = c("branch1" = "pink", "branch3" = "skyblue")) +  # Customcolors for branches
  stat_compare_means(method = "wilcox.test", label = "p.format", label.y = max(ponzniak_gsvaDF_merged$Melanocytic, na.rm = TRUE) * 1.05) +  # Add p-value
  theme_minimal() +
  labs(title = paste("Comparison of Melanocytic\n between branch1 and branch3"),
       y = "Melanocytic", x = "Branch") +
  theme(legend.position = "none")  # Hide legend for fill

  # Save the plot
print(p)
```



```{r}
# Create the boxplot and stripplot for each column
p <- ggplot(ponzniakDF_merged, aes(x = branch, y = Melanocytic, fill = branch)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +    # Boxplot without outliers
  geom_jitter(shape = 16, position = position_jitter(0.2), size = 3) +  # Stripplot
  scale_fill_manual(values = c("branch1" = "pink", "branch3" = "skyblue")) +  # Customcolors for branches
  stat_compare_means(method = "wilcox.test", label = "p.format", label.y = max(ponzniakDF_merged$Melanocytic, na.rm = TRUE) * 1.05) +  # Add p-value
  theme_minimal() +
  labs(title = paste("Comparison of Melanocytic\n between branch1 and branch3
                     based collaborator function"),
       y = "Melanocytic", x = "Branch") +
  theme(legend.position = "none")  # Hide legend for fill

  # Save the plot
print(p)
```


```{r, fig.height = 10, fig.width = 10}
mel_colors <- c(
  "MEL14" = "white",
  "MEL15" = "aquamarine4", 
  "MEL16" = "blue3", 
  "MEL17" = "deeppink",
  "MEL18" = "mistyrose2",
  "MEL19" = "brown2", 
  "MEL20" = "purple",
  "MEL21" = "khaki",
  "MEL23" = "palegreen", 
  "MEL24" = "steelblue", 
  "MEL25" = "tan",
  "MEL26" = "orange",
  "MEL27" = "yellow",
  "MEL29" = "seagreen3", 
  "MEL32" = "pink", 
  "MEL33" = "powderblue",
  "MEL34" = "cyan1",
  "MEL35" = "black", 
  "MEL37" = "grey", 
  "MEL38" = "lightcyan1", 
  "MEL39" = "lightpink3", 
  "MEL40" = "firebrick1", 
  "MEL42" = "magenta1", 
  "MEL44" = "bisque", 
  "MEL45" = "orange4", 
  "MEL47" = "gold1", 
  "MEL48" = "darkcyan", 
  "MEL49" = "darkslateblue", 
  "MEL50" = "deepskyblue3", 
  "MEL52" = "azure3", 
  "MEL53" = "peachpuff", 
  "MEL54" = "plum1", 
  "MEL55" = "royalblue", 
  "MEL56" = "lemonchiffon3", 
  "MEL58" = "tomato", 
  "MEL64" = "yellowgreen", 
  "MEL65" = "coral4", 
  "MEL68" = "salmon", 
  "MEL69" = "cornsilk"
)



breaks_list <- seq(-2, 2, by = 0.05)
print(length(breaks_list))
color_scale <- colorRampPalette(c("navy", "white", "red"))(length(breaks_list))

cl_cb <- function(hcl, mat){
    # Recalculate manhattan distances for reorder method
    dists <- dist(t(dgm_score_df), method = "manhattan")

    # Perform reordering according to OLO method
    hclust_olo <- reorder(hcl, dists)
    return(hclust_olo)
}

# Start SVG device
pdf("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/dgm_signature_analysis_heatmap_case.pdf", width = 10, height = 10)

pheatmap(dgm_score_df,
         scale = "row",
         border_color = "grey",
         show_colnames = F,
         cellheight = 10,
         cellwidth = 3,
         treeheight_row = 15, 
         treeheight_col = 25,
         color = color_scale, 
         breaks = breaks_list, 
         cluster_cols = TRUE, 
         cluster_rows = TRUE,
         #clustering_callback = cl_cb,
         annotation_col= invasive_pheno[, c("histopathology", "Immune_TV", "branch", "case")],
         annotation_colors = list(branch = c("branch1" = "tomato",
                                            "branch3" = "steelblue"),
                                  case = mel_colors),
         fontsize = 12,         # General font size
         fontsize_row = 10,     # Font size for row names
         fontsize_col = 10,   
         legend = T,
         annotation_legend = TRUE)

dev.off()
```
# boxplot
```{r}
dgm_scoreT <- as.data.frame(t(dgm_score_df))

#invasive_branch1

dgm_scoreT["branch"] <- "others"
dgm_scoreT[branch1_samples, "branch"] <- "branch1"
dgm_scoreT[branch3_samples, "branch"] <- "branch3"

dgm_scoreT$branch <- factor(dgm_scoreT$branch, levels = c("branch1", "branch3"))
table(dgm_scoreT$branch)

colnames(dgm_scoreT)

columns_to_plot <- names(dgm_scoreT)[!names(dgm_scoreT) == "branch"]
columns_to_plot
```


```{r}
library(ggpubr)

# Iterate through the columns and create boxplots and stripplots with p-value
for (column in columns_to_plot) {
  
  # Create the boxplot and stripplot for each column
  p <- ggplot(dgm_scoreT, aes(x = branch, y = dgm_scoreT[[column]], fill = branch)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +    # Boxplot without outliers
    geom_jitter(shape = 16, position = position_jitter(0.2), size = 3) +  # Stripplot
    scale_fill_manual(values = c("branch1" = "pink", "branch3" = "skyblue")) +  # Custom colors for branches
    stat_compare_means(method = "wilcox.test", label = "p.format", label.y = max(dgm_scoreT[[column]], na.rm = TRUE) * 1.05) +  # Add p-value
    theme_minimal() +
    labs(title = paste("Comparison of", column, "\n   between branch1 and branch3"),
         y = column, x = "Branch") +
    theme(legend.position = "none")  # Hide legend for fill

  # Save the plot
  print(p)
  ggsave(filename = paste0("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/Comparison_", column, ".pdf"), plot = p, width = 4, height = 6)
}
```



# extract the gene signatures for cybersort
```{r}
table(PCAI_merged_pheno$composition_TV)
# subset the immune ROIs from batch2
b2_immune_pheno <-PCAI_merged_pheno[(PCAI_merged_pheno$composition_TV == "immune") & (PCAI_merged_pheno$batch_str == "Batch_2"), ]
# create column named slide_rois
b2_immune_pheno$slide_rois <- paste0(b2_immune_pheno$slide.name, "_", b2_immune_pheno$roi_label)

# identify slide_rois associated with branch 1 and 3
tumor_branch1_rois <- invasive_pheno[invasive_pheno$branch == "branch1", "slide_rois"]
tumor_branch3_rois <- invasive_pheno[invasive_pheno$branch == "branch3", "slide_rois"]

b2_immune_pheno$branch <- "unknown"
b2_immune_pheno[b2_immune_pheno$slide_rois %in% tumor_branch1_rois, "branch"] <- "branch1"
b2_immune_pheno[b2_immune_pheno$slide_rois %in% tumor_branch3_rois, "branch"] <- "branch3"

table(b2_immune_pheno$branch)
# now manual match the unknown samples

unknown_samples <- b2_immune_pheno[b2_immune_pheno$branch == "unknown", ]
```



```{r}
write.csv(unknown_samples, "batch2_immune_ROIs_branch3_check_09302024.csv")
```


# check the batch2 data
```{r}
batch2_tumor_som <- load("PCAI_batch2_tumor_30x30_09082024.RData")
```

```{r}
# I need to initiate the datta first
data("opossom.genesets")
```





# now generate the normalized count data in comparison between branch1 and branch3
```{r, select signatures}
#library(oposSOM)
signature_names <- names(opossom.genesets)
#signature_names

selected_signatures <- c(
  "Landsberg_dedifferentiation_down", 
  "Joensson_Melanoma Pigmentation subtype", 
  "Harbst_melanoma_highgrade_up", 
  "Tirosh_Genes in the MITF program", 
  "Joensson_Melanoma Normal like subtype", 
  "TCGA_melanoma_keratin_high", 
  "Harbst_melanoma_lowgrade_up", 
  "Landsberg_dedifferentiation_up"
)

# double check the items'
missing_signatures <- selected_signatures[!selected_signatures %in% signature_names]
missing_signatures

```

```{r}

# Initialize a list to store converted genes
selected_genes_converted <- list()

# Loop through the selected signatures
for (signature_name in selected_signatures) {
  # Extract Ensembl gene IDs
  genes <- opossom.genesets[[signature_name]]$Genes
  selected_genes_converted[[signature_name]] <- genes
  }

# Checking the converted genes
selected_genes_converted
```


```{r}
library(EnsDb.Hsapiens.v79)
# Initialize list to store converted gene symbols
selected_genes_converted_symbols <- list()

# Loop through the selected signatures and convert to gene symbols
for (signature_name in selected_signatures) {
  # Extract the Ensembl gene IDs
  genes <- selected_genes_converted[[signature_name]]
  print(length(genes))
  converted_genes <- ensembldb::select(EnsDb.Hsapiens.v79, 
                                       keys= genes, 
                                       keytype = "GENEID", 
                                       columns = c("SYMBOL","GENEID"))$SYMBOL
  selected_genes_converted_symbols[[signature_name]] <- converted_genes 
  }
  
# Check the structure of the converted genes
str(selected_genes_converted_symbols)

# Print the first few entries for inspection
```


signature for N vs P
```{r}
NvsP_signatures <- c("AMIT_EGF_RESPONSE_40_HELA", 
                     "NAGASHIMA_EGF_SIGNALING_UP",
                     "REACTOME_DNA_METHYLATION",
                     "Reitmann_PA_MAPK-signaling", 
                     "AMIT_EGF_RESPONSE_40_MCF10A", 
                     "REACTOME_MEIOTIC_SYNAPSIS", 
                     "REACTOME_DNA_DAMAGE_TELOMERE_STRESS_INDUCED_SENESCENCE", 
                     "WANG_LSD1_TARGETS_DN", 
                     #"skeletal muscle cell differentiation", 
                     "PHONG_TNF_TARGETS_UP", 
                     "REACTOME_OXIDATIVE_STRESS_INDUCED_SENESCENCE", 
                     "VAQUERIZAS_Trachea_TF", 
                     "Tirosh_Genes in the MITF program", 
                     #"melanosome", 
                     "Tirosh_melanoma specific genes", 
                     "REACTOME_INNATE_IMMUNE_SYSTEM", 
                     "REACTOME_NEUTROPHIL_DEGRANULATION", 
                     "HALLMARK_INTERFERON_GAMMA_RESPONSE", 
                     #"neutrophil degranulation", 
                     "BROWNE_INTERFERON_RESPONSIVE_GENES", 
                     #"MHC class II protein complex binding", 
                     "Tirosh_genes preferentially expressed by Tregs", 
                     "KEGG_LYSOSOME" 
                     )




PvsMIS_signatures <- c("peptide antigen binding", 
                       "MHC class I protein complex", 
                       "ER to Golgi transport vesicle membrane", 
                       "Tirosh_Genes in the MITF program", 
                       "type I interferon signaling pathway", 
                       "melanosome", 
                       "phagocytic vesicle membrane", 
                       "mitochondrial inner membrane", 
                       "interferon-gamma-mediated signaling pathway", 
                       "Tirosh_melanoma specific genes", 
                       "antigen processing and presentation", 
                       "TAP1 binding", 
                       "ONKEN_UVEAL_MELANOMA_UP", 
                       "ENK_UV_RESPONSE_KERATINOCYTE_UP", 
                       "HALLMARK_MYC_TARGETS_V1", 
                       "WANG_TUMOR_INVASIVENESS_UP", 
                       "MHC class I peptide loading complex", 
                       "KEGG_OXIDATIVE_PHOSPHORYLATION",
                       "DER_IFN_ALPHA_RESPONSE_UP", 
                       "MHC class I protein binding", 
                       "RADAEVA_RESPONSE_TO_IFNA1_UP", 
                       "HALLMARK_OXIDATIVE_PHOSPHORYLATION", 
                       "REACTOME_CYCLIN_A_CDK2_ASSOCIATED_EVENTS_AT_S_PHASE_ENTRY", 
                       "G protein-coupled receptor activity", 
                       "extracellular matrix", 
                       "extracellular matrix organization", 
                       "extracellular region", 
                       "cell adhesion", 
                       "extracellular matrix structural constituent", 
                       "TCGA_melanoma_immune_high", 
                       "Tirosh_CAF-cell specific genes", 
                       "HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION", 
                       "Tirosh_expression higher in CAFs than in T-cells", 
                       "REACTOME_EXTRACELLULAR_MATRIX_ORGANIZATION", 
                       "MIKKELSEN_NPC_HCP_WITH_H3K27ME3", 
                       "Landsberg_dedifferentiation_up", 
                       "collagen-containing extracellular matrix", 
                       "REACTOME_COLLAGEN_CHAIN_TRIMERIZATION", 
                       "Tirosh_T-cell specific genes-melanoma", 
                       "inflammatory response")

# double check the items'
missing_signatures <- PvsMIS_signatures[!PvsMIS_signatures %in% signature_names]
PvsMIS_signatures_f <- PvsMIS_signatures[!PvsMIS_signatures %in% missing_signatures]

```

```{r}
# Convert Ensembl gene IDs to gene symbols for NvsP_signatures
NvsP_genes_converted <- setNames(lapply(NvsP_signatures, function(signature_name) {
  genes <- opossom.genesets[[signature_name]]$Genes
  
  # Convert Ensembl IDs to symbols
  converted_genes <- ensembldb::select(EnsDb.Hsapiens.v79, 
                                       keys = genes, 
                                       keytype = "GENEID", 
                                       columns = "SYMBOL")$SYMBOL
  
  return(converted_genes)
}), NvsP_signatures)  # Use signature names as list names

# Convert Ensembl gene IDs to gene symbols for PvsMIS_signatures
PvsMIS_genes_converted <- setNames(lapply(PvsMIS_signatures_f, function(signature_name) {
  genes <- opossom.genesets[[signature_name]]$Genes
  
  # Convert Ensembl IDs to symbols
  converted_genes <- ensembldb::select(EnsDb.Hsapiens.v79, 
                                       keys = genes, 
                                       keytype = "GENEID", 
                                       columns = "SYMBOL")$SYMBOL
  
  return(converted_genes)
}), PvsMIS_signatures_f)  # Use signature names as list names

# Print structure of the converted gene lists
str(NvsP_genes_converted)
str(PvsMIS_genes_converted)
```
run gsva


```{r}
table(b2_pheno_merged$histopathology)
tumor_raw_count_b2

# now creating DEG object 
b2_dds <-  DESeqDataSetFromMatrix(countData = round(tumor_raw_count_b2),
                                 colData = b2_pheno_merged,
                                 design = ~ histopathology)

# now create dds project
b2_dds <- DESeq(b2_dds)

# extract the normalized counts
b2_dds <- estimateSizeFactors(b2_dds)
b2_norm <- counts(b2_dds, normalize = T)
b2_log <- log2(b2_norm + 1)
b2_log_matrix <- as.matrix(b2_log)

```

# comparison between N, P and MIS
```{r}

N_P_pheno <- b2_pheno_merged[b2_pheno_merged$histopathology %in% c("N", "P"), ]
N_P_log_matrix <- b2_log_matrix[, rownames(N_P_pheno)]

P_MIS_pheno <- b2_pheno_merged[b2_pheno_merged$histopathology %in% c("P", "MIS"), ]
P_MIS_log_matrix <- b2_log_matrix[, rownames(P_MIS_pheno)]


NvsP_gsva <- gsva(ssgseaParam(N_P_log_matrix, NvsP_genes_converted), verbose = F)
PvsMIS_gsva <- gsva(ssgseaParam(P_MIS_log_matrix, PvsMIS_genes_converted), verbose = F)

NvsP_gsvaDF <- as.data.frame(t(NvsP_gsva))
PvsMIS_gsvaDF <- as.data.frame(t(PvsMIS_gsva))

NvsP_gsz <- as.data.frame(scale(NvsP_gsvaDF))
PvsMIS_gsz <- as.data.frame(scale(PvsMIS_gsvaDF))

# subset the data

```
```{r}
N_samples <- rownames(N_P_pheno[N_P_pheno$histopathology == "N", ])
P_samples <- rownames(N_P_pheno[N_P_pheno$histopathology == "P", ])

print(N_samples)
print(P_samples)

# assign the condition
NvsP_gsvaDF$condition <- "others"
NvsP_gsvaDF[rownames(NvsP_gsvaDF) %in% N_samples, "condition"] <- "N"
NvsP_gsvaDF[rownames(NvsP_gsvaDF) %in% P_samples, "condition"] <- "P"

table(NvsP_gsvaDF$condition)

```



volcano plot
```{r}
# Ensure to exclude the 'condition' column when applying the differential function
numeric_columns <- colnames(NvsP_gsvaDF)[!colnames(NvsP_gsvaDF) %in% "condition"]

# Function to calculate log fold change and p-value
calculate_differential <- function(data, pathway, group_col = "condition", group1 = "N", group2 = "P") {
  # Use dplyr::filter to avoid conflicts with other packages
  group1_data <- data %>% dplyr::filter(!!sym(group_col) == group1) %>% dplyr::select(pathway) %>% unlist()
  group2_data <- data %>% dplyr::filter(!!sym(group_col) == group2) %>% dplyr::select(pathway) %>% unlist()
  
  # Calculate log fold change and p-value
  if (length(group1_data) > 0 && length(group2_data) > 0) {
    log2FoldChange <- log2(mean(group2_data, na.rm = TRUE) / mean(group1_data, na.rm = TRUE))
    pvalue <- t.test(group1_data, group2_data)$p.value
    return(c(log2FoldChange, pvalue))
  } else {
    return(c(NA, NA))  # Handle cases where groups are empty or contain only NA
  }
}

# Apply the function across all numeric pathway columns
results <- sapply(numeric_columns, function(pathway) {
  calculate_differential(NvsP_gsvaDF, pathway)
})

# Transpose the results to create a proper dataframe
results_df <- as.data.frame(t(results))
colnames(results_df) <- c("log2FoldChange", "pvalue")

# Display the results
head(results_df)

results_df$padj <- p.adjust(results_df$pvalue, method = "BH")

# Create the volcano plot
ggplot(results_df, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point() +
  theme_minimal() +
  xlab("Log2 Fold Change") +
  ylab("-Log10 Adjusted P-value") +
  geom_hline(yintercept = -log10(0.05), col = "red", linetype = "dashed") +  # Significance threshold
  geom_vline(xintercept = c(-1, 1), col = "blue", linetype = "dashed")  # LogFC threshold
```

```{r, fig.height= 8, fig.width = 10}
library(EnhancedVolcano)


results_clean <- na.omit(results_df)


# Specify the filename for the SVG file
#output_filename <- "/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/B2_NvsP_DGEplot.pdf"
 

# Start SVG device
#pdf(output_filename, width = 7, height = 8) # Set appropriate width and height

EnhancedVolcano(results_clean,
    lab = rownames(results_clean),
    title = 'Normal vs Precusor',
    #selectLab = c('PMEL','CDK2', 
     #             "CDHR1", "CDH1", "ITGA4"),
    x = 'log2FoldChange',
    y = 'pvalue',
    labSize = 5.0, 
    pointSize = 3.0, 
    FCcutoff = 1,
    pCutoff = 0.05,
    subtitle = bquote('FDR cut off 0.001 and absolute LFC cut off 1'))

#dev.off()
```




```{r}
pathway <- "Tirosh_Genes in the MITF program"

# Q-Q plot for the group "N"
qqnorm(NvsP_gsvaDF[NvsP_gsvaDF$condition == "N", pathway], main = "Q-Q plot for Group N", col = "blue")
qqline(NvsP_gsvaDF[NvsP_gsvaDF$condition == "N", pathway], col = "blue")

# Q-Q plot for the group "P"
qqnorm(NvsP_gsvaDF[NvsP_gsvaDF$condition == "P", pathway], main = "Q-Q plot for Group P", col = "red")
qqline(NvsP_gsvaDF[NvsP_gsvaDF$condition == "P", pathway], col = "red")

# Alternatively, you can plot both groups on the same Q-Q plot for comparison
qqplot(
  NvsP_gsvaDF[NvsP_gsvaDF$condition == "N", pathway],
  NvsP_gsvaDF[NvsP_gsvaDF$condition == "P", pathway],
  main = "Q-Q plot comparing Group N and Group P"
)
abline(0, 1, col = "black")  # Add a reference line
```
```{r, fig.height = 10, fig.width = 10}
# Load required packages
library(ggplot2)
library(reshape2)

# Assuming your data is in 'NvsP_gsvaDF' with pathways as columns
# Reshape the data to long format for easier plotting
gsva_long <- melt(NvsP_gsvaDF, id.vars = "condition", variable.name = "Tirosh_Genes in the MITF program", value.name = "gsva_score")

# Generate a faceted scatter plot to show GSVA scores across pathways
ggplot(gsva_long, aes(x = condition, y = gsva_score)) +
  geom_boxplot() +
  facet_wrap(~ `Tirosh_Genes in the MITF program`, scales = "free_y") +  # Facet by pathway
  theme_minimal() +
  xlab("Condition") +
  ylab("GSVA Score") +
  theme(strip.text = element_text(size = 10))
```


```{r figure 2 N vs P, fig.height= 6, fig.width= 12}
breaks_list <- seq(-2, 2, by = 0.05)
print(length(breaks_list))
color_scale <- colorRampPalette(c("navy", "white", "red"))(length(breaks_list))

# Start SVG device
pdf("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/NvsP_signature_analysis_heatmap.pdf", width = 12, height = 6)

cl_cb <- function(hcl, mat){
    # Recalculate manhattan distances for reorder method
    dists <- dist(NvsP_gsz, method = "manhattan")

    # Perform reordering according to OLO method
    hclust_olo <- reorder(hcl, dists)
    return(hclust_olo)
}

pheatmap(t(NvsP_gsz),
         scale = "row",
         border_color = "grey",
         show_colnames = F,
         cellheight = 10,
         cellwidth = 3,
         treeheight_row = 15, 
         treeheight_col = 25,
         color = color_scale, 
         breaks = breaks_list, 
         cluster_cols = TRUE, 
         cluster_rows = FALSE,
         clustering_callback = cl_cb,
         annotation_col= N_P_pheno[#"localization", 
                                     "histopathology" 
                                     #"MISvsInvasive", "Immune_TV", 
                                     #"slide.name"
                                     ],
         annotation_colors = list(histopathology = c("N" = "steelblue",
                                                     "P" = "forestgreen", 
                                                     "MIS" = "pink", 
                                                     "RGP" = "orange", 
                                                     "VGP" = "tomato")), 
          #                        case = mel_colors),
         fontsize = 12,         # General font size
         fontsize_row = 10,     # Font size for row names
         fontsize_col = 10,   
         legend = T,
         annotation_legend = TRUE)


# Manually render the heatmap using grid.draw
#grid::grid.draw(pheatmap_result$gtable)

# Close the PDF device
dev.off()
```


```{r figure 2 P vs MIS, fig.height= 6, fig.width= 12}
breaks_list <- seq(-2, 2, by = 0.05)
print(length(breaks_list))
color_scale <- colorRampPalette(c("navy", "white", "red"))(length(breaks_list))

# Start SVG device
pdf("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/PvsMIS_signature_analysis_heatmap.pdf", width = 12, height = 6)

cl_cb <- function(hcl, mat){
    # Recalculate manhattan distances for reorder method
    dists <- dist(PvsMIS_gsz, method = "manhattan")

    # Perform reordering according to OLO method
    hclust_olo <- reorder(hcl, dists)
    return(hclust_olo)
}

pheatmap(t(PvsMIS_gsz),
         scale = "row",
         border_color = "grey",
         show_colnames = F,
         cellheight = 10,
         cellwidth = 3,
         treeheight_row = 15, 
         treeheight_col = 25,
         color = color_scale, 
         breaks = breaks_list, 
         cluster_cols = TRUE, 
         cluster_rows = FALSE,
         clustering_callback = cl_cb,
         annotation_col= P_MIS_pheno[#"localization", 
                                     "histopathology" 
                                     #"MISvsInvasive", "Immune_TV", 
                                     #"slide.name"
                                     ],
         annotation_colors = list(histopathology = c("N" = "steelblue",
                                                     "P" = "forestgreen", 
                                                     "MIS" = "pink", 
                                                     "RGP" = "orange", 
                                                     "VGP" = "tomato")), 
          #                        case = mel_colors),
         fontsize = 12,         # General font size
         fontsize_row = 10,     # Font size for row names
         fontsize_col = 10,   
         legend = T,
         annotation_legend = TRUE)


# Manually render the heatmap using grid.draw
#grid::grid.draw(pheatmap_result$gtable)

# Close the PDF device
dev.off()
```

# now gsea 
```{r}
b2_gsva <- gsva(ssgseaParam(b2_log_matrix, selected_genes_converted_symbols), verbose = F)

b2_gsvaDF <- as.data.frame(t(b2_gsva))

b2_gsvaDF

b2_gsz <- as.data.frame(scale(b2_gsvaDF))
```

```{r}
b2_logT <- as.data.frame(t(b2_log))
b2_logT

selected_genes <- c("MLANA", "SOX9", "SOX10", "PMEL", "MITF", "NGFR", "AXL")

b2_sub_log <- b2_logT[, selected_genes]

b2_sub_log


b2_sub_log$ID <- rownames(b2_sub_log)
b2_gsz$ID <- rownames(b2_gsz)

# Merge based on the "ID" column
merged_df <- merge(b2_sub_log, b2_gsz, by = "ID", all = TRUE)

# Restore rownames if needed
rownames(merged_df) <- merged_df$ID

merged_df$ID <- NULL  # Opti

merged_df
```
```{r}
colnames(merged_df)
```


```{r}
library(ggpubr)

my_data <- merged_df
# Define the selected columns
selected_columns <- c("MLANA", "SOX9", "SOX10", "PMEL", "MITF", "NGFR", "AXL")
matched_columns <- c("Landsberg_dedifferentiation_down", 
                    "Joensson_Melanoma Pigmentation subtype", 
                    "Harbst_melanoma_highgrade_up", 
                    "Tirosh_Genes in the MITF program", 
                    "Joensson_Melanoma Normal like subtype",
                    "TCGA_melanoma_keratin_high", "Harbst_melanoma_lowgrade_up", "Landsberg_dedifferentiation_up")

# Define the output PDF file
output_pdf <- "/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/PCA_signature_analysis_heatmap_with_log.pdf"

# Open the PDF device
pdf(output_pdf, width = 7, height = 5)

# Loop through each selected column
for (x_col in selected_columns) {
  # Loop through all other columns in the dataframe
  for (y_col in setdiff(matched_columns, x_col)) {
    # Use backticks around column names to handle spaces
    p <- ggplot(my_data, aes(x = .data[[x_col]], y = .data[[y_col]])) +
      geom_point(alpha = 0.7) +
      theme_minimal() +
      labs(title = paste("Scatter plot of", x_col, "vs", y_col),
           x = x_col,
           y = y_col) +
      geom_smooth(method = "lm", se = FALSE, color = "blue") +
      stat_cor(method = "pearson", label.x.npc = "left", label.y.npc = "top")

    # Print the plot to the PDF
    print(p)
  }
}

# Close the PDF device
dev.off()

cat("Scatter plots with correlation and p-values saved to", output_pdf, "\n")
```

```{r}
#merged_seg <- inner_join(tumor_s1_seg_deathDF, immune_s1_seg_deathDF, by = "ROILabel")

# collected
pairs(my_data)
```

```{r, fig.height= 8, fig.width= 10, size = 12}
library(ggcorrplot)
library(corrplot)


merged_corr <- round(cor(merged_df), 1)

pro.p.mat <- cor_pmat(merged_corr)

pro_corr2 <- replace(merged_corr, is.na(merged_corr), 0)
pro.p.mat2 <- replace(pro.p.mat, is.na(pro.p.mat), 0)


col <- colorRampPalette(c("#4477AA", "#77AADD", "#FFFFFF", "#EE9988", "#BB4444"))

pdf("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/correlation_plot.pdf", width = 10, height = 10)

corrplot(pro_corr2, 
         #method = "color", 
         col = col(100),  
         #type = "upper", 
         order = "hclust", 
         #addrect = 5,
         #rect.col = "red",
         addCoef.col = "black", # Add coefficient of correlation
         tl.col = "darkblue", tl.srt = 45, #Text label color and rotation
         # Combine with significance level
         p.mat = pro.p.mat2, sig.level = 0.05,
         #insig = "blank",
         # hide correlation coefficient on the principal diagonal
         diag = TRUE)

dev.off()

cat("Correlation plot saved to correlation_plot.pdf\n")
```

# generate the pheatmap
```{r figure 2 sub, fig.height= 6, fig.width= 12}
breaks_list <- seq(-2, 2, by = 0.05)
print(length(breaks_list))
color_scale <- colorRampPalette(c("navy", "white", "red"))(length(breaks_list))

# Start SVG device
pdf("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/PCA_signature_analysis_heatmap_with_log.pdf", width = 12, height = 6)

cl_cb <- function(hcl, mat){
    # Recalculate manhattan distances for reorder method
    dists <- dist(b2_gsz, method = "manhattan")

    # Perform reordering according to OLO method
    hclust_olo <- reorder(hcl, dists)
    return(hclust_olo)
}

pheatmap(t(b2_gsz),
         scale = "row",
         #border_color = "grey",
         show_colnames = F,
         cellheight = 10,
         cellwidth = 1,
         treeheight_row = 15, 
         treeheight_col = 25,
         color = color_scale, 
         breaks = breaks_list, 
         cluster_cols = TRUE, 
         cluster_rows = FALSE,
         clustering_callback = cl_cb,
         annotation_col= b2_pheno_merged[#"localization", 
                                     "histopathology" 
                                     #"MISvsInvasive", "Immune_TV", 
                                     #"slide.name"
                                     ],
         annotation_colors = list(histopathology = c("N" = "steelblue",
                                                     "P" = "forestgreen", 
                                                     "MIS" = "pink", 
                                                     "RGP" = "orange", 
                                                     "VGP" = "tomato")), 
          #                        case = mel_colors),
         fontsize = 12,         # General font size
         fontsize_row = 10,     # Font size for row names
         fontsize_col = 10,   
         legend = T,
         annotation_legend = TRUE)


# Manually render the heatmap using grid.draw
#grid::grid.draw(pheatmap_result$gtable)

# Close the PDF device
dev.off()
```

# scatter plot
```{r figure 2 sub, fig.height= 6, fig.width= 12}
breaks_list <- seq(-2, 2, by = 0.05)
print(length(breaks_list))
color_scale <- colorRampPalette(c("navy", "white", "red"))(length(breaks_list))

# Start SVG device
pdf("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/PCA_signature_analysis_heatmap_with_log.pdf", width = 12, height = 6)

cl_cb <- function(hcl, mat){
    # Recalculate manhattan distances for reorder method
    dists <- dist(merged_df, method = "manhattan")

    # Perform reordering according to OLO method
    hclust_olo <- reorder(hcl, dists)
    return(hclust_olo)
}

pheatmap(t(merged_df),
         scale = "row",
         #border_color = "grey",
         show_colnames = F,
         cellheight = 10,
         cellwidth = 1,
         treeheight_row = 15, 
         treeheight_col = 25,
         color = color_scale, 
         breaks = breaks_list, 
         cluster_cols = TRUE, 
         cluster_rows = FALSE,
         clustering_callback = cl_cb,
         annotation_col= b2_pheno_merged[#"localization", 
                                     "histopathology" 
                                     #"MISvsInvasive", "Immune_TV", 
                                     #"slide.name"
                                     ],
         annotation_colors = list(histopathology = c("N" = "steelblue",
                                                     "P" = "forestgreen", 
                                                     "MIS" = "pink", 
                                                     "RGP" = "orange", 
                                                     "VGP" = "tomato")), 
          #                        case = mel_colors),
         fontsize = 12,         # General font size
         fontsize_row = 10,     # Font size for row names
         fontsize_col = 10,   
         legend = T,
         annotation_legend = TRUE)


# Manually render the heatmap using grid.draw
#grid::grid.draw(pheatmap_result$gtable)

# Close the PDF device
dev.off()
```






# run stage 2 data
```{r}
library(ComplexHeatmap)

# Create the heatmap with specific legend parameters
ht <- Heatmap(t(b2_gsz),
              #name = "matrix_28",
              col = color_scale,
              clustering_distance_columns = "manhattan",
              clustering_callback = cl_cb,
              show_column_names = FALSE,
              row_names_gp = gpar(fontsize = 10),
              column_names_gp = gpar(fontsize = 10),
              top_annotation = HeatmapAnnotation(
                  histopathology = b2_pheno_merged["histopathology"],
                  col = list(histopathology = c("N" = "steelblue",
                                                "P" = "forestgreen", 
                                                "MIS" = "pink", 
                                                "RGP" = "orange", 
                                                "VGP" = "tomato"))
              ),
              heatmap_legend_param = list(
                  title = "Expression",
                  at = seq(-4, 4, 2),  # Control breaks in the legend
                  legend_height = unit(5, "cm")  # Adjust legend size
              )
)

# Adjust layout by using the `draw` function
draw(ht, heatmap_legend_side = "right", annotation_legend_side = "right")
```


```{r}
library(ComplexHeatmap)
library(dendextend)
library(seriation)
library(circlize)
```


```{r, fig.height = 10, fig.width= 10}
# Define the color scale
color_scale <- colorRamp2(c(-2, 0, 2), c("blue", "white", "red"))

# Create the row annotation based on histopathology (since samples are rows)
row_annotation <- rowAnnotation(
  histopathology = b2_pheno_merged$histopathology,
  col = list(histopathology = c(
    "N" = "steelblue",
    "P" = "forestgreen", 
    "MIS" = "pink", 
    "RGP" = "orange", 
    "VGP" = "tomato"))
)

# Clustering callback function for optimal leaf ordering (OLO)
cl_cb <- function(hcl, mat) {
  dists <- dist(mat, method = "manhattan")  # Manhattan distance
  hclust_olo <- reorder(hcl, dists)  # Optimal leaf ordering
  return(hclust_olo)
}

# Perform hierarchical clustering on the rows (samples) with Manhattan distance
dist_cols <- dist(b2_gsz, method = "manhattan")  # Distance between samples (rows)
hclust_cols <- hclust(dist_cols, method = "complete")
hclust_cols_olo <- cl_cb(hclust_cols, b2_gsz)  # Optimal leaf ordering for samples

# Build the heatmap using ComplexHeatmap
Heatmap(
  b2_gsz,  # Data (samples as rows, signatures as columns)
  name = "GSZ Score",  # Legend title
  col = color_scale,  # Color scale
  cluster_rows = as.dendrogram(hclust_cols_olo),  # Optimal leaf ordering for rows (samples)
  cluster_columns = FALSE,  # Do not cluster columns (signatures)
  show_row_names = FALSE, # Show row names (samples)
  show_column_names = TRUE,  # Show column names (gene signatures)
  row_dend_width = unit(15, "mm"),  # Dendrogram width for rows
  left_annotation = row_annotation,  # Add the row annotation (samples)
  heatmap_legend_param = list(title = "GSZ Score", title_gp = gpar(fontsize = 12), labels_gp = gpar(fontsize = 10)),
  row_names_gp = gpar(fontsize = 10),  # Font size for row names (samples)
  column_names_gp = gpar(fontsize = 10),  # Font size for column names (signatures)
  height = unit(15, "cm"),  # Adjust the overall height of the heatmap
  width = unit(10, "cm")  # Adjust the overall width of the heatmap
)

```


```{r immune branch analysis}
immune_pheno <- b2_immune_pheno[b2_immune_pheno$branch %in% c("branch1", "branch3"), ]
immune_raw <- PCAI_merged_raw[, rownames(immune_pheno)]
immune_raw

# perform dds for this
# now creating DEG object 
immune_dds <-  DESeqDataSetFromMatrix(countData = round(immune_raw),
                                 colData = immune_pheno,
                                 design = ~ branch)

immune_dds <- DESeq(immune_dds)
immune_vsd <- vst(immune_dds, blind = FALSE)
# comparison
immune_res <- results(immune_dds, contrast = c("branch", "branch1", "branch3"))
immune_res <- immune_res[order(immune_res$padj), ]
immune_sig_genes <- immune_res[which(immune_res$padj <= 0.05), ]

b1_up_genes <- rownames(immune_sig_genes[immune_sig_genes$log2FoldChange > 0, ])
b3_up_genes <- rownames(immune_sig_genes[immune_sig_genes$log2FoldChange < 0, ])

# identify the number of up and downregulated genes
cat("Number of upregulated genes in branch1:", length(b1_up_genes), "\n")
cat("Number of downregulated genes in branch1 (upregulated in branch3):", length(b3_up_genes), "\n")
```
```{r}
convert_to_entrez <- function(gene_list) {
    gene_conversion <- bitr(gene_list, fromType = "SYMBOL",
                            toType = "ENTREZID",
                            OrgDb = org.Hs.eg.db)
    return(unique(gene_conversion$ENTREZID))
}

# Convert both gene sets
up_genes_entrez <- convert_to_entrez(b1_up_genes)
down_genes_entrez <- convert_to_entrez(b3_up_genes)

# Check for conversion success
cat("Upregulated genes converted to Entrez IDs:", length(up_genes_entrez), "\n")
cat("Downregulated genes converted to Entrez IDs:", length(down_genes_entrez), "\n")
```
# define the pathways
```{r, use MSigDB C7}
# Load MSigDB C7 gene sets
c7_gene_sets <- msigdbr(species = "Homo sapiens", category = "C7")
c7_list <- split(c7_gene_sets$entrez_gene, c7_gene_sets$gs_name)  # Use Entrez IDs

# Convert c7_list to TERM2GENE format (a data frame)
c7_df <- data.frame(
    TERM = rep(names(c7_list), times = sapply(c7_list, length)),
    GENE = unlist(c7_list)
)

# Perform enrichment for upregulated genes
c7_enrichment_up <- enricher(up_genes_entrez, TERM2GENE = c7_df, pAdjustMethod = "BH", qvalueCutoff = 0.05)
# Perform enrichment for downregulated genes
c7_enrichment_down <- enricher(down_genes_entrez, TERM2GENE = c7_df, pAdjustMethod = "BH", qvalueCutoff = 0.05)
```

```{r, use MSigDB C2}
# Load MSigDB C7 gene sets
c2_gene_sets <- msigdbr(species = "Homo sapiens", category = "C2")
c2_list <- split(c2_gene_sets$entrez_gene, c2_gene_sets$gs_name)  # Use Entrez IDs

# Convert c7_list to TERM2GENE format (a data frame)
c2_df <- data.frame(
    TERM = rep(names(c2_list), times = sapply(c2_list, length)),
    GENE = unlist(c2_list)
)

# Perform enrichment for upregulated genes
c2_enrichment_up <- enricher(up_genes_entrez, TERM2GENE = c2_df, pAdjustMethod = "BH", qvalueCutoff = 0.05)
# Perform enrichment for downregulated genes
c2_enrichment_down <- enricher(down_genes_entrez, TERM2GENE = c2_df, pAdjustMethod = "BH", qvalueCutoff = 0.05)
```



```{r, KEGG pathway enrichment}
# Perform KEGG enrichment for upregulated genes
kegg_enrichment_up <- enrichKEGG(gene = up_genes_entrez,
                                 organism = 'hsa',
                                 pvalueCutoff = 0.05,
                                 pAdjustMethod = "BH")

# Perform KEGG enrichment for downregulated genes
kegg_enrichment_down <- enrichKEGG(gene = down_genes_entrez,
                                   organism = 'hsa',
                                   pvalueCutoff = 0.05,
                                   pAdjustMethod = "BH")
```

```{r, rectome pathway enrichement}
# Perform Reactome enrichment for upregulated genes
reactome_enrichment_up <- enrichPathway(gene = up_genes_entrez,
                                       organism = "human",
                                       pvalueCutoff = 0.05,
                                       pAdjustMethod = "BH")

# Perform Reactome enrichment for downregulated genes
reactome_enrichment_down <- enrichPathway(gene = down_genes_entrez,
                                         organism = "human",
                                         pvalueCutoff = 0.05,
                                         pAdjustMethod = "BH")
```

```{r, hallmark enrichment}
library(ReactomePA)
hallmark_gene_sets <- msigdbr(species = "Homo sapiens", category = "H")

hallmark_list <- split(hallmark_gene_sets$entrez_gene, hallmark_gene_sets$gs_name)

hallmark_df <- data.frame(
    TERM = rep(names(hallmark_list), times = sapply(hallmark_list, length)),
    GENE = unlist(hallmark_list)
)

# Perform enrichment for upregulated genes
hallmark_enrichment_up <- enricher(up_genes_entrez, TERM2GENE = hallmark_df, pAdjustMethod = "BH", qvalueCutoff = 0.05)

# Perform enrichment for downregulated genes
hallmark_enrichment_down <- enricher(down_genes_entrez, TERM2GENE = hallmark_df, pAdjustMethod = "BH", qvalueCutoff = 0.05)

```


```{r, fig.height = 6, fig.width= 6}
# Set the output file to save the plots as a high-resolution PDF
pdf("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/Enrichment_Analysis_Plots.pdf", width=8, height=8)  # Set the resolution and size of the PDF

# Upregulated in branch1 - C7
if (!is.null(c7_enrichment_up) && nrow(c7_enrichment_up) > 0) {
    dotplot(c7_enrichment_up, showCategory=12, title="C7 Immunologic Signatures Enriched in Branch1")
} else {
    print("No significant C7 enrichment found for upregulated genes in branch1.")
}

# Downregulated in branch1 (Enriched in branch3) - C7
if (!is.null(c7_enrichment_down) && nrow(c7_enrichment_down) > 0) {
    dotplot(c7_enrichment_down, showCategory=12, title="C7 Immunologic Signatures Enriched in Branch3")
} else {
    print("No significant C7 enrichment found for downregulated genes in branch1.")
}

# Upregulated in branch1 - C2
if (!is.null(c2_enrichment_up) && nrow(c2_enrichment_up) > 0) {
    dotplot(c2_enrichment_up, showCategory=12, title="C2 Immunologic Signatures Enriched in Branch1")
} else {
    print("No significant C2 enrichment found for upregulated genes in branch1.")
}

# Downregulated in branch1 (Enriched in branch3) - C2
if (!is.null(c2_enrichment_down) && nrow(c2_enrichment_down) > 0) {
    dotplot(c2_enrichment_down, showCategory=12, title="C2 Immunologic Signatures Enriched in Branch3")
} else {
    print("No significant C2 enrichment found for downregulated genes in branch1.")
}

# Upregulated in branch1 - KEGG
if (!is.null(kegg_enrichment_up) && nrow(kegg_enrichment_up) > 0) {
    dotplot(kegg_enrichment_up, showCategory=12, title="KEGG Pathways Enriched in Branch1")
} else {
    print("No significant KEGG enrichment found for upregulated genes in branch1.")
}

# Downregulated in branch1 (Enriched in branch3) - KEGG
if (!is.null(kegg_enrichment_down) && nrow(kegg_enrichment_down) > 0) {
    dotplot(kegg_enrichment_down, showCategory=12, title="KEGG Pathways Enriched in Branch3")
} else {
    print("No significant KEGG enrichment found for downregulated genes in branch1.")
}

# Upregulated in branch1 - Reactome
if (!is.null(reactome_enrichment_up) && nrow(reactome_enrichment_up) > 0) {
    dotplot(reactome_enrichment_up, showCategory=12, title="Reactome Pathways Enriched in Branch1")
} else {
    print("No significant Reactome enrichment found for upregulated genes in branch1.")
}

# Downregulated in branch1 (Enriched in branch3) - Reactome
if (!is.null(reactome_enrichment_down) && nrow(reactome_enrichment_down) > 0) {
    dotplot(reactome_enrichment_down, showCategory=12, title="Reactome Pathways Enriched in Branch3")
} else {
    print("No significant Reactome enrichment found for downregulated genes in branch1.")
}

# Upregulated in branch1 - Hallmark
if (!is.null(hallmark_enrichment_up) && nrow(hallmark_enrichment_up) > 0) {
    dotplot(hallmark_enrichment_up, showCategory=12, title="Hallmark Immune Pathways Enriched in Branch1")
} else {
    print("No significant Hallmark enrichment found for upregulated genes in branch1.")
}

# Downregulated in branch1 (Enriched in branch3) - Hallmark
if (!is.null(hallmark_enrichment_down) && nrow(hallmark_enrichment_down) > 0) {
    dotplot(hallmark_enrichment_down, showCategory=12, title="Hallmark Immune Pathways Enriched in Branch3")
} else {
    print("No significant Hallmark enrichment found for downregulated genes in branch1.")
}

# Close the PDF device to save the file
dev.off()
```
# now import paper based pathways particularly for immun
```{r}
# Convert the list to TERM2GENE format
TERM2GENE <- data.frame(
    TERM = rep(names(immune_geneSig_list), sapply(immune_geneSig_list, length)),
    GENE = unlist(immune_geneSig_list)
)

# Inspect the TERM2GENE data frame
str(TERM2GENE)

immune_enrichment_up <- enricher(up_genes_entrez, TERM2GENE = TERM2GENE, pAdjustMethod = "BH", qvalueCutoff = 0.05)
immune_enrichment_down <- enricher(down_genes_entrez, TERM2GENE = TERM2GENE, pAdjustMethod = "BH", qvalueCutoff = 0.05)

# no gene mapped

```

```{r}
immune_norm_counts <- counts(immune_dds, normalized = T)
immune_log <- log2(immune_norm_counts + 1)

immune_log_matrix <- as.matrix(immune_log)

Immune_gsva <- gsva(ssgseaParam(immune_log_matrix, immune_geneSig_list), verbose = F)

Immune_gsvaDF <- as.data.frame(t(Immune_gsva))

Immune_gsvaDF
#IO_gsvaDF$PATIENT_ID <- rownames(IO_gsvaDF)
```


```{r, fig.height= 10, fig.width= 10}
breaks_list <- seq(-2, 2, by = 0.05)
print(length(breaks_list))
color_scale <- colorRampPalette(c("navy", "white", "red"))(length(breaks_list))

# Start SVG device
pdf("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/Immune_Analysis_Plots.pdf", width = 8, height = 8)

pheatmap(t(Immune_gsvaDF),
         scale = "row",
         border_color = "grey",
         show_colnames = F,
         cellheight = 10,
         cellwidth = 3,
         treeheight_row = 15, 
         treeheight_col = 15,
         color = color_scale, 
         breaks = breaks_list, 
         annotation_col= immune_pheno[, c(#"localization", 
                                     "histopathology", 
                                     #"MISvsInvasive", "Immune_TV", 
                                     "slide.name", "branch")],
         annotation_colors = list(branch = c("branch1" = "tomato",
                                                     "branch3" = "skyblue")), 
          #                        case = mel_colors),
         fontsize = 12,         # General font size
         fontsize_row = 10,     # Font size for row names
         fontsize_col = 10,   
         legend = T)

# Manually render the heatmap using grid.draw
#grid::grid.draw(pheatmap_result$gtable)

# Close the PDF device
dev.off()


print(pheatmap_result)
```

```{r}
immune_branch1_samples <- rownames(immune_pheno[immune_pheno$branch == "branch1", ])
immune_branch3_samples <- rownames(immune_pheno[immune_pheno$branch == "branch3", ])
```


# now do boxplot for all of them
```{r}
Immune_gsvaDF$branch <- "unknown"
Immune_gsvaDF[rownames(Immune_gsvaDF) %in% immune_branch1_samples, "branch"] <- "branch1"
Immune_gsvaDF[rownames(Immune_gsvaDF) %in% immune_branch3_samples, "branch"] <- "branch3"

table(Immune_gsvaDF$branch)
```

```{r}
library(ggpubr)

# Assuming your data is in a data frame called `data`
# Ensure that "branch" is a factor
Immune_gsvaDF$branch <- factor(Immune_gsvaDF$branch, levels = c("branch1", "branch3"))

# List of columns for comparison
columns_to_plot <- c("Antitumor cytokines", "B cells", "Cancer-associated fibroblasts", 
                     "Checkpoint molecules", "Co-stimulatory ligands", "Co-stimulatory receptors", 
                     "Effector cell traffic", "Effector cells", "Endothelium", "Granulocyte traffic", 
                     "Immune Suppression by Myeloid Cells", "M1 signature", "Macrophage and DC traffic", 
                     "Matrix", "Matrix remodeling", "MHCI", "MHCII", "Myeloid cells traffic", 
                     "Neutrophil signature", "NK cells", "Protumor cytokines", "T cells", 
                     "Th1 signature", "Th2 signature", "Treg", "Treg and Th2 traffic", 
                     "Tumor proliferation rate", "Tumor-associated Macrophages")

# Iterate through the columns and create boxplots and stripplots with p-value
for (column in columns_to_plot) {
  
  # Create the boxplot and stripplot for each column
  p <- ggplot(Immune_gsvaDF, aes(x = branch, y = Immune_gsvaDF[[column]], fill = branch)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +    # Boxplot without outliers
    geom_jitter(shape = 16, position = position_jitter(0.2), size = 3) +  # Stripplot
    scale_fill_manual(values = c("branch1" = "pink", "branch3" = "skyblue")) +  # Custom colors for branches
    stat_compare_means(method = "wilcox.test", label = "p.format", label.y = max(Immune_gsvaDF[[column]], na.rm = TRUE) * 1.05) +  # Add p-value
    theme_minimal() +
    labs(title = paste("Comparison of", column, "\n   between branch1 and branch3"),
         y = column, x = "Branch") +
    theme(legend.position = "none")  # Hide legend for fill

  # Save the plot
  print(p)
  ggsave(filename = paste0("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/Comparison_", column, ".pdf"), plot = p, width = 4, height = 6)
}
```


# access the TCGA data for primary and mets comparison
```{r}
setwd("/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/skcm_tcga_gdc/")

skcm_tpm <- read.csv("data_mrna_seq_tpm.txt", sep = "\t", row.names = 1)
head(skcm_tpm)

# Convert Entrez Gene IDs to Gene Symbols
entrez_ids <- rownames(skcm_tpm)
gene_symbols <- mapIds(org.Hs.eg.db, keys = entrez_ids, column = "SYMBOL", keytype = "ENTREZID", multiVals = "first")

# 1. Check if the lengths of the Entrez IDs and gene_symbols match
if(length(rownames(skcm_tpm)) != length(gene_symbols)) {
  stop("The length of Entrez IDs and gene symbols do not match!")
}

# 2. Create a data frame with Entrez IDs, Gene Symbols, and skcm_tpm data
skcm_tpm_with_symbols <- data.frame(Entrez_ID = rownames(skcm_tpm), Gene_Symbol = gene_symbols, skcm_tpm)

# 3. Remove rows with NA gene symbols
skcm_tpm_filtered <- na.omit(skcm_tpm_with_symbols)

# 4. Assign Gene_Symbol as row names and remove the 'Entrez_ID' and 'Gene_Symbol' columns
rownames(skcm_tpm_filtered) <- skcm_tpm_filtered$Gene_Symbol
skcm_tpm_filtered <- skcm_tpm_filtered[ , !(names(skcm_tpm_filtered) %in% c("Entrez_ID", "Gene_Symbol"))]

# 5. Check the updated dataset
# 1. Move the rownames (gene symbols) into a new column named 'Gene'
skcm_tpm_filtered$Gene <- rownames(skcm_tpm_filtered)

# 2. Remove the rownames
rownames(skcm_tpm_filtered) <- NULL

# 3. Reorder the columns so that 'Gene' is the first column
skcm_tpm_filtered <- skcm_tpm_filtered[, c("Gene", colnames(skcm_tpm_filtered)[colnames(skcm_tpm_filtered) != "Gene"])]

# 4. Write the data frame to a .txt file
write.table(skcm_tpm_filtered, file = "skcm_tpm_filtered.txt", sep = "\t", row.names = FALSE, quote = FALSE)
```
# now read skcm clinical data


```{r}
setwd("/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/skcm_tcga_gdc/")
skcm_clinical <- read.csv("data_clinical_sample.txt", sep = "\t", comment.char = "#")
skcm_clinical

```



# access the skcm data for NR and R comparison
```{r import and organized the data}
# set the working directory
setwd("/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/published_projects/Project6_IO/publication_prep/")
# load the RNA and clinical dataset
# Load the data
IO_tpm <- read.csv("PublicData/mel_dfci_2019/data_RNA_Seq_expression_tpm.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)
# Drop the 'Entrez_Gene_Id' column
IO_tpm <- IO_tpm[, !names(IO_tpm) %in% "Entrez_Gene_Id"]

duplicated_hugo <- duplicated(IO_tpm$Hugo_Symbol)
# 1-Mar and 2-Mar
# Modify only the duplicated Hugo_Symbol entries by appending "_2"
IO_tpm$Hugo_Symbol[duplicated_hugo] <- paste0(IO_tpm$Hugo_Symbol[duplicated_hugo], "_2")

# Set the Hugo_Symbol as row names now that duplicates are resolved
rownames(IO_tpm) <- IO_tpm$Hugo_Symbol

# Optionally, drop the Hugo_Symbol column if it's now set as rownames
IO_tpm <- IO_tpm[, !names(IO_tpm) %in% "Hugo_Symbol"]


clinical_data <- read.csv("PublicData/mel_dfci_2019/data_clinical_patient.txt", sep = "\t", comment.char = "#")
dfci_purity_data <- read.csv("PublicData/mel_dfci_2019/data_clinical_sample.txt", sep = "\t", comment.char = "#")

dfci_clinical_merged <- merge(clinical_data, dfci_purity_data, by = "PATIENT_ID")

dfci_clinical_merged
```

```{r}
# Ensure that both selected_samples and IO_tpm column names are characters with no extra spaces
# Check for samples in selected_samples that are not in IO_tpm column names
missing_samples <- setdiff(selected_samples, colnames(IO_tpm))
missing_samples
```


```{r}
# only select skin samples
dfci_skin_purity <- dfci_clinical_merged[dfci_clinical_merged$PRIMARY_TYPE == "skin", ]

skin_samples <- dfci_skin_purity$SAMPLE_ID

skin_samples <- trimws(as.character(skin_samples))
colnames(IO_tpm) <- trimws(as.character(colnames(IO_tpm)))

matched_samples <- intersect(skin_samples, colnames(IO_tpm))

print(length(matched_samples))

matched_clinical <- dfci_skin_purity[dfci_skin_purity$SAMPLE_ID %in% matched_samples, ]

# Data clean and extract the features
matched_clinical<- matched_clinical %>% 
  dplyr::filter(BR %in% c("Complete Response", "Partial Response", "Progressive Disease")) %>% 
  mutate(response = if_else(BR == "Progressive Disease", "NR", "R")) %>% 
  mutate(EVENT = if_else(PFS_STATUS == "1:PROGRESSION", 1, 0)) %>% 
  dplyr::select("PATIENT_ID", "SAMPLE_ID", "response", "BR", "PFS_STATUS", "EVENT", "PFS_MONTHS", "PURITY")

matched_skin_tpm <- IO_tpm[, matched_clinical$SAMPLE_ID]

gene_names <- rownames(matched_skin_tpm)


numeric_IO_skin_tpm <- as.data.frame(lapply(matched_skin_tpm, function(x) as.numeric(as.character(x))))

str(numeric_IO_skin_tpm)

numeric_IO_skin_tpm
# If you have gene names in the first column that you want to keep as rownames:
rownames(numeric_IO_skin_tpm) <- gene_names
# Apply log2 transformation (log2(TPM + 1)) to the numeric data
log_IO_skin_tpm <- log2(numeric_IO_skin_tpm + 1)
# View the transformed data
head(log_IO_skin_tpm)

```



```{r}
# Adjust expression data by regressing out the purity for each gene
adjusted_tpm <- log_IO_skin_tpm   # Create a copy for adjusted data

for (gene in rownames(log_IO_skin_tpm)) {
  # Regress out the purity from each gene's expression
  fit <- lm(as.numeric(log_IO_skin_tpm[gene, ]) ~ matched_clinical$PURITY)
  residuals <- resid(fit)
  
  # Replace the original expression values with the residuals (purity-adjusted values)
  adjusted_tpm[gene, ] <- residuals
}

```





```{r}
IO_branch_gsva<- gsva(ssgseaParam(as.matrix(adjusted_tpm), gene_signatures), verbose = F)
IO_branch_gsvaDF <- as.data.frame(t(IO_branch_gsva))

IO_branch_gsvaDF$SAMPLE_ID <- rownames(IO_branch_gsvaDF)

IO_branch_gsvaDF

```

```{r}
matched_clinical_merged <- merge(matched_clinical, IO_branch_gsvaDF, by = "SAMPLE_ID")
table(matched_clinical_merged$response)

```


```{r, fig.height= 4, fig.width= 3}
# Create the boxplot and stripplot for BRANCH1_SIGNATURE
branch1 <- ggplot(matched_clinical_merged, aes(x = response, y = BRANCH1_SIGNATURE, fill = response)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +    # Boxplot without outliers
    geom_jitter(shape = 16, position = position_jitter(0.2), size = 1) +  # Stripplot
    scale_fill_manual(values = c("NR" = "pink", "R" = "skyblue")) +  # Custom colors for branches
    stat_compare_means(method = "wilcox.test", label = "p.format", label.y = max(matched_clinical_merged$BRANCH1_SIGNATURE, na.rm = TRUE) * 1.05) +  # Add p-value
    labs(title = "Comparison of BRANCH1 \nNR vs R",
         y = "BRANCH1_SIGNATURE", x = "Tumor Type") +  # Adjust title and axis labels
    theme(
        panel.background = element_rect(fill = "white", color = NA),  # White background
        plot.background = element_rect(fill = "white", color = NA),   # White plot background
        panel.grid.major = element_blank(),                           # Remove major gridlines
        panel.grid.minor = element_blank(),                           # Remove minor gridlines
        axis.line = element_line(color = "black"),                    # Show x and y axis lines
        axis.ticks = element_line(color = "black"),                   # Show axis ticks
        axis.text = element_text(color = "black"),                    # Ensure axis text is black
        legend.position = "none"                                      # Hide legend for fill
    )


branch3 <- ggplot(matched_clinical_merged, aes(x = response, y = BRANCH3_SIGNATURE, fill = response)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +    # Boxplot without outliers
    geom_jitter(shape = 16, position = position_jitter(0.2), size = 1) +  # Stripplot
    scale_fill_manual(values = c("NR" = "pink", "R" = "skyblue")) +  # Custom colors for branches
    stat_compare_means(method = "wilcox.test", label = "p.format", label.y = max(matched_clinical_merged$BRANCH3_SIGNATURE, na.rm = TRUE) * 1.05) +  # Add p-value
    labs(title = "Comparison of BRANCH3 \nNR vs R",
         y = "BRANCH3_SIGNATURE", x = "Tumor Type") +  # Adjust title and axis labels
    theme(
        panel.background = element_rect(fill = "white", color = NA),  # White background
        plot.background = element_rect(fill = "white", color = NA),   # White plot background
        panel.grid.major = element_blank(),                           # Remove major gridlines
        panel.grid.minor = element_blank(),                           # Remove minor gridlines
        axis.line = element_line(color = "black"),                    # Show x and y axis lines
        axis.ticks = element_line(color = "black"),                   # Show axis ticks
        axis.text = element_text(color = "black"),                    # Ensure axis text is black
        legend.position = "none"                                      # Hide legend for fill
    )

ggsave("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/branch1_NRvsR_purity_adjusted.pdf", plot = branch1, width = 3, height = 4, device = "pdf", dpi = 300)

ggsave("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/branch3_NRvsR_purity_adjusted.pdf", plot = branch3, width = 3, height = 4, device = "pdf", dpi = 300)


# Print the plot
print(branch1)
print(branch3)
```

```{r}
# 1. Transpose the data
cibersort_dfci2019 <- as.data.frame(t(numeric_tTMP))

# 2. Add the gene names as a new column
cibersort_dfci2019$Gene <- rownames(cibersort_dfci2019)

# 3. Reorder the columns so that 'Gene' is the first column
cibersort_dfci2019 <- cibersort_dfci2019[, c("Gene", colnames(cibersort_dfci2019)[colnames(cibersort_dfci2019) != "Gene"])]

# 4. Write to a txt file in the correct format for CIBERSORT
write.table(cibersort_dfci2019, file = "dfci_tpm_cibersort_input.txt", sep = "\t", row.names = FALSE, quote = FALSE)
```

```{r}
table(skcm_clinical$SAMPLE_TYPE)
rownames(skcm_clinical) <- gsub("-", ".", skcm_clinical$SAMPLE_ID)

# Check the rownames
head(rownames(skcm_clinical))

# Subset primary solid tumor samples
primary_tumor <- rownames(skcm_clinical[skcm_clinical$SAMPLE_TYPE == "Primary solid Tumor", ])

# Subset metastatic tumor samples
mets_tumor <- rownames(skcm_clinical[skcm_clinical$SAMPLE_TYPE == "Metastatic", ])
```

# read the absolute result
```{r, skcm}
setwd("/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/processed_data/")
skcm_cibersort <- read.csv("CIBERSORTx_skcm_branch_09302024.csv")[, c("Mixture", "branch1", "branch3")]
skcm_ciber_f <- skcm_cibersort[skcm_cibersort$Mixture %in% c(primary_tumor, mets_tumor), ]
skcm_ciber_f$mets <- "unknown"
skcm_ciber_f[skcm_ciber_f$Mixture %in% primary_tumor, "mets"] <- "primary"
skcm_ciber_f[skcm_ciber_f$Mixture %in% mets_tumor, "mets"] <- "metastatic"

table(skcm_ciber_f$mets)
```
decon looks really bad, now simply do signatue analysis
```{r}

head(skcm_tpm_filtered)
rownames(skcm_tpm_filtered) <- skcm_tpm_filtered$Gene
skcm_tpm_filtered$Gene <- NULL

log_skcm_tpm <- log2(skcm_tpm_filtered + 1) 
log_skcm_tpm
```



```{r}
tpm_core_ids <- sub("(TCGA.[^.]+.[^.]+).[A-Z0-9]+", "\\1", colnames(skcm_tpm_filtered))
colnames(skcm_tpm_filtered) <- tpm_core_ids

```



```{r}
setwd("/Users/yshi797/Dropbox (Partners HealthCare)/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/processed_data/")
skcm_purity_data <- read.csv("mat_TCGA.tsv", sep = "\t")
# Extract the core ID from the `purity_sample_barcodes`
# Use a simpler approach to extract only the part starting with 'TCGA' and ending with the third part of the code.
# Extract core IDs from purity data
purity_core_ids <- sapply(skcm_purity_data$Tumor_Sample_Barcode, function(x) {
  # Remove "SKCM-" prefix
  x <- sub("^SKCM-", "", x)
  # Split by hyphens
  parts <- strsplit(x, "-")[[1]]
  # Extract the first three parts (TCGA, XX, XXXX)
  core_id <- paste(parts[1:3], collapse = ".")
  return(core_id)
})

# Print the first few purity core IDs
cat("First few Purity Core IDs:\n")
print(head(purity_core_ids))


```

```{r}
length(tpm_core_ids)
```


```{r}
skcm_purity_data$core_id <- purity_core_ids
skcm_purity_data[, c("Tumor_Sample_Barcode", "core_id")]

skcm_purity_data_f <- skcm_purity_data[skcm_purity_data$core_id %in% tpm_core_ids, ]

skcm_tpm_match <- skcm_tpm_filtered[, skcm_purity_data_f$core_id]

skcm_purity_data_f
```



```{r}
log_tpm <- log2(skcm_tpm_match + 1)
skcm_purity_data_f<- skcm_purity_data_f[match(colnames(skcm_tpm_match), skcm_purity_data_f$core_id),]
skcm_purity_data_f

```

```{r}
# Adjust expression data by regressing out the purity for each gene
adjusted_tpm <- log_tpm  # Create a copy for adjusted data

for (gene in rownames(log_tpm)) {
  # Regress out the purity from each gene's expression
  fit <- lm(as.numeric(log_tpm[gene, ]) ~ purity_df$purity)
  residuals <- resid(fit)
  
  # Replace the original expression values with the residuals (purity-adjusted values)
  adjusted_tpm[gene, ] <- residuals
}
```




```{r}
# Find matching core IDs
matching_core_ids <- intersect(tpm_core_ids, purity_core_ids)
num_matches <- length(matching_core_ids)

# Display the number of matches and the matching IDs
cat("Number of matching core IDs:", num_matches, "\n")
cat("Matching Core IDs:\n")
print(matching_core_ids)
```



boxplot comparison
```{r, branch1 absolute}
ggbetweenstats(data = skcm_ciber_f,
               x = mets,
               y = branch3,
               messages = F,
               bf.message = F,
               mean.size = 2,
               mean.color = "black",
               ylab = "branch1 signature",
               pairwise.comparisons = FALSE,
               results.subtile = FALSE,
               centrality.label.args = list(size = 4),
               ggsignif.args = list(textsize = 4, tip_length = 0.01)) + 
                theme(axis.title.y.right = element_blank(), 
                          axis.text.y.right = element_blank(), 
                          axis.ticks.y.right = element_blank(),
                          text = element_text(size = 12),       # Increase overall font size
                          axis.title = element_text(size = 12), # Increase axis titles font size
                          axis.text = element_text(size = 12))
                          #geom_signif(comparisons = list(c("Tumor", "Immune")), test.args = list(exact = FALSE)    
```

```{r}
skcm_gsva <- gsva(ssgseaParam(as.matrix(log_skcm_tpm), gene_signatures), verbose = F)
skcm_gsvaDF <- as.data.frame(t(skcm_gsva))

skcm_gsvaDF$mets <- "unknown"
skcm_gsvaDF[rownames(skcm_gsvaDF) %in% primary_tumor, "mets"] <- "primary"
skcm_gsvaDF[rownames(skcm_gsvaDF) %in% mets_tumor, "mets"] <- "metastatic"
skcm_gsvaDF <- skcm_gsvaDF[skcm_gsvaDF$mets != "unknown", ]

table(skcm_gsvaDF$mets)

```


```{r, fig.height= 4, fig.width= 3}
# Create the boxplot and stripplot for BRANCH1_SIGNATURE
branch1 <- ggplot(skcm_gsvaDF, aes(x = mets, y = BRANCH1_SIGNATURE, fill = mets)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +    # Boxplot without outliers
    geom_jitter(shape = 16, position = position_jitter(0.2), size = 1) +  # Stripplot
    scale_fill_manual(values = c("metastatic" = "pink", "primary" = "skyblue")) +  # Custom colors for branches
    stat_compare_means(method = "wilcox.test", label = "p.format", label.y = max(skcm_gsvaDF$BRANCH1_SIGNATURE, na.rm = TRUE) * 1.05) +  # Add p-value
    labs(title = "Comparison of BRANCH1 \nPrimary vs Metastatic Tumors",
         y = "BRANCH1_SIGNATURE", x = "Tumor Type") +  # Adjust title and axis labels
    theme(
        panel.background = element_rect(fill = "white", color = NA),  # White background
        plot.background = element_rect(fill = "white", color = NA),   # White plot background
        panel.grid.major = element_blank(),                           # Remove major gridlines
        panel.grid.minor = element_blank(),                           # Remove minor gridlines
        axis.line = element_line(color = "black"),                    # Show x and y axis lines
        axis.ticks = element_line(color = "black"),                   # Show axis ticks
        axis.text = element_text(color = "black"),                    # Ensure axis text is black
        legend.position = "none"                                      # Hide legend for fill
    )


branch3 <- ggplot(skcm_gsvaDF, aes(x = mets, y = BRANCH3_SIGNATURE, fill = mets)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +    # Boxplot without outliers
    geom_jitter(shape = 16, position = position_jitter(0.2), size = 1) +  # Stripplot
    scale_fill_manual(values = c("metastatic" = "pink", "primary" = "skyblue")) +  # Custom colors for branches
    stat_compare_means(method = "wilcox.test", label = "p.format", label.y = max(skcm_gsvaDF$BRANCH3_SIGNATURE, na.rm = TRUE) * 1.05) +  # Add p-value
    labs(title = "Comparison of BRANCH3 \nPrimary vs Metastatic Tumors",
         y = "BRANCH3_SIGNATURE", x = "Tumor Type") +  # Adjust title and axis labels
    theme(
        panel.background = element_rect(fill = "white", color = NA),  # White background
        plot.background = element_rect(fill = "white", color = NA),   # White plot background
        panel.grid.major = element_blank(),                           # Remove major gridlines
        panel.grid.minor = element_blank(),                           # Remove minor gridlines
        axis.line = element_line(color = "black"),                    # Show x and y axis lines
        axis.ticks = element_line(color = "black"),                   # Show axis ticks
        axis.text = element_text(color = "black"),                    # Ensure axis text is black
        legend.position = "none"                                      # Hide legend for fill
    )

ggsave("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/branch1_PvsM.pdf", plot = branch1, width = 3, height = 4, device = "pdf", dpi = 300)

ggsave("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/branch3_PvsM.pdf", plot = branch3, width = 3, height = 4, device = "pdf", dpi = 300)


# Print the plot
print(branch1)
print(branch3)
```

do GSVA calculation
```{r}
IO_gsva <- gsva(ssgseaParam(as.matrix(t(log_tTMP)), gene_signatures), verbose = F)
IO_gsvaDF <- as.data.frame(t(IO_gsva))
IO_gsvaDF$PATIENT_ID <- rownames(IO_gsvaDF)
```




```{r}
# Data merging
mergedDF <- merge(meta_data, IO_gsvaDF, by = "PATIENT_ID", all = FALSE)
colnames(mergedDF)[1] <- "Patient_ID"
rownames(mergedDF) <- mergedDF$Patient_ID

mergedDF
```


```{r}
# Data clean and extract the features
mergedDF <- mergedDF %>% 
  filter(BR %in% c("Complete Response", "Partial Response", "Progressive Disease")) %>% 
  mutate(response = if_else(BR == "Progressive Disease", "NR", "R")) %>% 
  mutate(EVENT = if_else(PFS_STATUS == "1:PROGRESSION", 1, 0)) %>% 
  dplyr::select("Patient_ID", "response", "BR", "PFS_STATUS", "EVENT", "PFS_MONTHS", "BRANCH1_SIGNATURE", "BRANCH3_SIGNATURE") 

mergedDF
```


```{r nr vs r, fig.height= 4, fig.width= 3}
# Create the boxplot and stripplot for BRANCH1_SIGNATURE
dfci_branch1 <- ggplot(mergedDF, aes(x = response, y = BRANCH1_SIGNATURE, fill = response)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +    # Boxplot without outliers
    geom_jitter(shape = 16, position = position_jitter(0.2), size = 1) +  # Stripplot
    scale_fill_manual(values = c("NR" = "orange", "R" = "lightgreen")) +  # Custom colors for branches
    stat_compare_means(method = "wilcox.test", label = "p.format", label.y = max(mergedDF$BRANCH1_SIGNATURE, na.rm = TRUE) * 1.05) +  # Add p-value
    labs(title = "Comparison of BRANCH1 \nNR vs R",
         y = "BRANCH1_SIGNATURE", x = "Tumor Type") +  # Adjust title and axis labels
    theme(
        panel.background = element_rect(fill = "white", color = NA),  # White background
        plot.background = element_rect(fill = "white", color = NA),   # White plot background
        panel.grid.major = element_blank(),                           # Remove major gridlines
        panel.grid.minor = element_blank(),                           # Remove minor gridlines
        axis.line = element_line(color = "black"),                    # Show x and y axis lines
        axis.ticks = element_line(color = "black"),                   # Show axis ticks
        axis.text = element_text(color = "black"),                    # Ensure axis text is black
        legend.position = "none"                                      # Hide legend for fill
    )


dfci_branch3 <- ggplot(mergedDF, aes(x = response, y = BRANCH3_SIGNATURE, fill = response)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +    # Boxplot without outliers
    geom_jitter(shape = 16, position = position_jitter(0.2), size = 1) +  # Stripplot
    scale_fill_manual(values = c("NR" = "orange", "R" = "lightgreen")) +  # Custom colors for branches
    stat_compare_means(method = "wilcox.test", label = "p.format", label.y = max(mergedDF$BRANCH3_SIGNATURE, na.rm = TRUE) * 1.05) +  # Add p-value
    labs(title = "Comparison of BRANCH3 \nNR vs R",
         y = "BRANCH3_SIGNATURE", x = "Tumor Type") +  # Adjust title and axis labels
    theme(
        panel.background = element_rect(fill = "white", color = NA),  # White background
        plot.background = element_rect(fill = "white", color = NA),   # White plot background
        panel.grid.major = element_blank(),                           # Remove major gridlines
        panel.grid.minor = element_blank(),                           # Remove minor gridlines
        axis.line = element_line(color = "black"),                    # Show x and y axis lines
        axis.ticks = element_line(color = "black"),                   # Show axis ticks
        axis.text = element_text(color = "black"),                    # Ensure axis text is black
        legend.position = "none"                                      # Hide legend for fill
    )


ggsave("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/branch1_NvsNR.pdf", plot = dfci_branch1, width = 3, height = 4, device = "pdf", dpi = 300)

ggsave("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/branch3_NvsNR.pdf", plot = dfci_branch3, width = 3, height = 4, device = "pdf", dpi = 300)

# Print the plot
print(dfci_branch1)
print(dfci_branch3)


```



# extract the count
```{r}
# output the normalized and log-transformed data
invasive_dds <- estimateSizeFactors(invasive_dds)
invasive_norm <- counts(invasive_dds, normalize = T)
invasive_log <- log2(invasive_norm + 1)

invasive_log
```

# now look the expression difference between genes
```{r}
selected_genes <- c("SOX10","MITF","MLANA", "CDH1", "MKI67", "RPS6", "PMEL", "AXL", "NGFR")

invasive_logT <- as.data.frame(t(invasive_log[selected_genes, ]))

invasive_logT$sample_label <- rownames(invasive_logT)

invasive_logT

invasive_merged_log <- merge(invasive_pheno, invasive_logT, by = "sample_label")

invasive_merged_log

graphDF <- invasive_merged_log[, c("branch", selected_genes)]

graphDF
```

```{r branch gene comparison, fig.width = 4, fig.height = 6}
# Iterate through the columns and create boxplots and stripplots with p-value
for (gene in selected_genes) {
  
  # calculate sample size per group (branch1 and 3)
  n_branch1 <- sum(graphDF$branch == "branch1")
  n_branch3 <- sum(graphDF$branch == "branch3")
  
   # Create the boxplot and stripplot for each column
  p <- ggplot(graphDF, aes(x = branch, y = graphDF[[gene]], fill = branch)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +    # Boxplot without outliers
    geom_jitter(shape = 16, position = position_jitter(0.2), size = 3) +  # Stripplot
    scale_fill_manual(values = c("branch1" = "pink", "branch3" = "skyblue")) +  # Custom colors for branches
    stat_compare_means(method = "wilcox.test", label = "p.format", label.y = max(graphDF[[gene]], na.rm = TRUE) * 1.05) +  # Add p-value
    theme_minimal() +
    labs(title = paste("Comparison of", gene, "\nbetween branch1 and branch3"),
         y = paste("log-transformed", gene), x = "Branch") +
    scale_x_discrete(labels = c(paste0("branch1\n(n=", n_branch1, ")"),
                                paste0("branch3\n(n=", n_branch3, ")"))) +  # Add sample size to x-ticks
    theme(legend.position = "none")  # Hide legend for fill

  # Save the plot
  print(p)
  ggsave(filename = paste0("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/Branch_comparison_", gene, ".pdf"), plot = p, width = 4, height = 6)
}
```













# cytokine and chemokine reference
# https://www.immport.org/shared/genelists
```{r}
setwd("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/")
chemokines <- read.csv("signatures/Chemokines.txt", sep = "\t")
cytokines <- read.csv("signatures/Cytokines.txt", sep = "\t")
chemokines_rep <- read.csv("signatures/Chemokine_Receptors.txt", sep = "\t")
cytokines_rep <- read.csv("signatures/Cytokine_Receptors.txt", sep = "\t")
interleukins <- read.csv("signatures/Interleukins.txt", sep = "\t")
interleukin_rep <- read.csv("signatures/Interleukins_Receptors.txt", sep = "\t")
APP <- read.csv("signatures/Antigen_Processing_and_Presentation.txt", sep = "\t")
interferons <- read.csv("signatures/Interferons.txt", sep = "\t")
TNF <- read.csv("signatures/TNF_Family_Members.txt", sep = "\t")
TGF <- read.csv("signatures/TGF-b_Family_Members.txt", sep = "\t")

```




```{r}
invasive_logT = as.data.frame(t(invasive_log))
invasive_logT

# Add sample_label column to invasive_logT
invasive_logT$sample_label <- rownames(invasive_logT)

# Merge both dataframes based on sample_label

# Display the first few rows of the merged dataframe
```

```{r, fig.height= 6,fig.width = 4}
library(ggpubr)
combined_df <- merge(invasive_pheno, invasive_logT, by = "sample_label")

# Calculate sample size for each branch
sample_sizes <- combined_df %>%
  group_by(branch) %>%
  summarise(n = n()) %>%
  mutate(branch_label = paste0(branch, " (n=", n, ")"))

# Merge sample size labels into the dataset
combined_df <- combined_df %>%
  left_join(sample_sizes, by = "branch")

# Create the boxplot with statistical test
ggplot(combined_df, aes(x = branch_label, y = A2M, fill = branch)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +  # Boxplot with transparency
  geom_jitter(width = 0.2, size = 2, alpha = 0.6) +  # Dots for individual points
  stat_compare_means(method = "wilcox.test", label = "p.signif") +  # Wilcoxon test
  theme_minimal() +
  labs(title = "Comparison of A2M expression between Branch 1 and Branch 3",
       x = "Branch",
       y = "A2M Expression") +
  scale_fill_manual(values = c("branch1" = "steelblue", "branch3" = "tomato")) +
  theme(legend.position = "none")
```


```{r}
setwd("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/")

generate_boxplot <- function(gene_list, combined_df, file_name, height = 6, width = 4, p_threshold = 0.05) {
  # If the gene list contains two columns (Symbol and Synonyms), combine and extract unique genes
  if ("Symbol" %in% colnames(gene_list) & "Synonyms" %in% colnames(gene_list)) {
    gene_list$combined <- paste(gene_list$Symbol, gene_list$Synonyms, sep = ", ")
    gene_list$gene_set <- strsplit(gene_list$combined, ", ")
    genes <- unique(unlist(gene_list$gene_set))
  } else {
    # If the gene list is a single column (e.g., cytokines)
    genes <- unique(gene_list[[1]])  # Extract first column if it's a dataframe
  }
  
  # Find common genes between gene list and column names of combined_df
  common_genes <- intersect(colnames(combined_df), genes)
  
  significant_genes <- c()  # Store significant genes
  
  for (gene in common_genes) {
    # Perform Wilcoxon test
    p_value <- wilcox.test(combined_df[[gene]] ~ combined_df$branch)$p.value
    
    # Proceed only if the result is significant
    if (p_value < p_threshold) {
      significant_genes <- c(significant_genes, gene)  # Store significant gene
      
      # Create the boxplot
      boxplot <- ggplot(combined_df, aes(x = branch, y = .data[[gene]], fill = branch)) +
        geom_boxplot(alpha = 0.5, outlier.shape = NA) +  # Boxplot with transparency
        geom_jitter(width = 0.2, size = 2, alpha = 0.6) +  # Dots for individual points
        stat_compare_means(method = "wilcox.test", label = "p.signif") +  # Wilcoxon test
        theme_minimal() +
        labs(title = paste(gene, "expression Branch 1 vs Branch 3"),
             x = "Branch",
             y = paste(gene, "Expression")) +
        scale_fill_manual(values = c("branch1" = "steelblue", "branch3" = "tomato")) +
        theme(legend.position = "none")
      
      # Save the boxplot as a PDF
      ggsave(paste0(file_name, "_", gene, ".pdf"), plot = boxplot, width = width, height = height, device = "pdf", dpi = 300)
      
      # Also display the boxplot
      print(boxplot)
    }
  }
  
  # Return significant genes
  return(significant_genes)
}

#generate_boxplot(chemokines, combined_df, "chemokines")
generate_boxplot(cytokines, combined_df, "cytokines")

```


```{r}
generate_boxplot(cytokines_rep, combined_df, "cytokines_rep")
generate_boxplot(chemokines_rep, combined_df, "chemokines_rep")
generate_boxplot(interleukins, combined_df, "interleukins")
generate_boxplot(interleukin_rep, combined_df, "interleukin_rep")
generate_boxplot(APP, combined_df, "APP")
generate_boxplot(interferons, combined_df, "interferons")
generate_boxplot(TNF, combined_df, "TNF")
generate_boxplot(TGF, combined_df, "TGF")


```


```{r}
chemokines_rep <- read.csv("signatures/Chemokine_Receptors.txt", sep = "\t")
cytokines_rep <- read.csv("signatures/Cytokine_Receptors.txt", sep = "\t")
interleukins <- read.csv("signatures/Interleukins.txt", sep = "\t")
interleukin_rep <- read.csv("signatures/Interleukins_Receptors.txt", sep = "\t")
APP <- read.csv("signatures/Antigen_Processing_and_Presentation.txt", sep = "\t")
interferons <- read.csv("signatures/Interferons.txt", sep = "\t")
TNF <- read.csv("signatures/TNF_Family_Members.txt", sep = "\t")
TGF <- read.csv("signatures/TGF-b_Family_Members.txt", sep = "\t")

```





# create a function to generate this plot
```{r}
setwd("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/")

# Define a function to generate heatmaps for gene sets with scaling and sample selection options
generate_heatmap <- function(gene_list, invasive_log, invasive_pheno, heatmap_title, file_name, do_scale = TRUE, n_samples = NULL, height = 10, width = 10) {
  
  # If the gene list contains two columns (Symbol and Synonyms), combine and extract unique genes
  if ("Symbol" %in% colnames(gene_list) & "Synonyms" %in% colnames(gene_list)) {
    gene_list$combined <- paste(gene_list$Symbol, gene_list$Synonyms, sep = ", ")
    gene_list$gene_set <- strsplit(gene_list$combined, ", ")
    genes <- unique(unlist(gene_list$gene_set, ", "))
  } else {
    # If the gene list is a single column (e.g., cytokines)
    genes <- unique(gene_list)
  }
  
  # Find common genes between gene list and the rownames of invasive_log
  common_genes <- intersect(rownames(invasive_log), genes)
  
  # Subset the data
  gene_log <- invasive_log[c(common_genes, "MLANA"), ]
  gene_log_sorted <- gene_log[, order(gene_log["MLANA", ])]
  
  # If `n_samples` is specified, select the top and bottom n_samples based on MLANA expression
  if (!is.null(n_samples)) {
    top_n_samples <- colnames(gene_log_sorted)[1:n_samples]
    bottom_n_samples <- colnames(gene_log_sorted)[(ncol(gene_log_sorted) - n_samples + 1):ncol(gene_log_sorted)]
    selected_samples <- c(top_n_samples, bottom_n_samples)
    gene_log_sorted <- gene_log_sorted[, selected_samples]  # Subset to top and bottom samples
  }
  
  # Subset gene expression matrix (without MLANA) and extract MLANA values for the barplot
  gene_heatmap <- gene_log_sorted[common_genes, ]
  mlana_expression <- gene_log_sorted["MLANA", ]
  
  # Check if we should scale the data or not
  if (do_scale) {
    # Rescale data by rows (genes) using Z-score normalization (row-wise scaling)
    gene_heatmap_scaled <- t(apply(gene_heatmap, 1, scale))
  } else {
    # No scaling, use the raw data
    gene_heatmap_scaled <- gene_heatmap
  }
  
  # Perform hierarchical clustering on rows and apply optimal leaf ordering (OLO)
  dist_rows <- dist(gene_heatmap_scaled)
  hc_rows <- hclust(dist_rows)
  dend <- as.dendrogram(hc_rows)
  row_order <- seriate(dist_rows, method = "OLO")  # Apply OLO
  ordered_dend <- rotate(dend, get_order(row_order))  # Reorder dendrogram with OLO using `rotate()`
  
  # Sort annotations based on the columns of gene_log_sorted
  annotation_df_sorted <- invasive_pheno[colnames(gene_log_sorted), ]
  
  # Create the top annotation with MLANA, histopathology, and branch
  top_annotation <- HeatmapAnnotation(
    bar = anno_barplot(mlana_expression, border = FALSE, gp = gpar(fill = "white")),
    histopathology = annotation_df_sorted$histopathology, # Add histopathology annotation
    branch = annotation_df_sorted$branch,                 # Add branch annotation
    annotation_name_side = "left",
    col = list(
      histopathology = c("VGP" = "red", "RGP" = "orange"),  # Customize colors for histopathology
      branch = c("branch1" = "forestgreen", "branch3" = "purple")  # Customize colors for branch
    )
  )
  
  # Generate the heatmap with clustering of rows and ordered columns based on MLANA
  heatmap <- Heatmap(
    as.matrix(gene_heatmap_scaled), 
    name = "Expression", 
    top_annotation = top_annotation, 
    cluster_rows = ordered_dend, 
    cluster_columns = FALSE,  # Don't cluster columns, they are already ordered by MLANA
    show_column_names = FALSE, 
    show_row_names = TRUE, 
    column_title = heatmap_title,
    heatmap_legend_param = list(title = "Expression Level"),
    border = TRUE,
    rect_gp = gpar(col = "grey") # Add grey borders around each box
  )
  
  # Capture the ComplexHeatmap plot as a grid object
  heatmap_grob <- grid.grabExpr(draw(heatmap))
  
  # Save the heatmap using ggsave (needs conversion to a grid object)
  ggsave(file_name, plot = heatmap_grob, width = width, height = height, device = "pdf", dpi = 300)
  
  # Also display the heatmap in the R environment
  draw(heatmap)
}

# Apply the function to different gene lists
# Use `do_scale = TRUE` for scaling, `do_scale = FALSE` if you don't want scaling
# The `n_samples = 10` ensures we select the top 10 and bottom 10 samples, use `n_samples = NULL` to use all samples
generate_heatmap(chemokines, invasive_log, invasive_pheno, "Chemokines Expression Heatmap", "figures/Chemokines_Expression_Heatmap_top10.pdf", do_scale = TRUE, n_samples = 10, height = 10, width = 8)
generate_heatmap(cytokines, invasive_log, invasive_pheno, "Cytokines Expression Heatmap", "figures/Cytokines_Expression_Heatmap_top10.pdf", do_scale = TRUE, n_samples = 10, height = 25, width = 8)
generate_heatmap(chemokines_rep, invasive_log, invasive_pheno, "Chemokine Receptors Expression Heatmap", "figures/Chemokine_Receptors_Expression_Heatmap_top10.pdf", do_scale = T, n_samples = 10,  height = 10, width = 8) # Use all samples
generate_heatmap(cytokines_rep, invasive_log, invasive_pheno, "Cytokine Receptors Expression Heatmap", "figures/Cytokine_Receptors_Expression_Heatmap_top10.pdf", do_scale = T, n_samples = 10, height = 25, width = 8)  # Use all samples, without scaling
generate_heatmap(interleukins, invasive_log, invasive_pheno, "Interleukins Expression Heatmap", "figures/Interleukins_Expression_Heatmap_top10.pdf", do_scale = TRUE, n_samples = 10, height = 10, width = 8)

generate_heatmap(interleukin_rep, invasive_log, invasive_pheno, "Interleukins Expression Heatmap", "figures/Interleukins_Receptors_Expression_Heatmap_top10.pdf", do_scale = TRUE, n_samples = 10, height = 10, width = 8)

generate_heatmap(APP, invasive_log, invasive_pheno, "Antigen Presentation Expression Heatmap", "figures/Antigen_Presentation_Expression_Heatmap_top10.pdf", do_scale = TRUE, n_samples = 10, height = 20, width = 8)

generate_heatmap(interferons, invasive_log, invasive_pheno, "Interferons Expression Heatmap", "figures/Interferons_Expression_Heatmap_top10.pdf", do_scale = TRUE, n_samples = 10, height = 4, width = 8)

generate_heatmap(TNF, invasive_log, invasive_pheno, "TNF-Alpha Expression Heatmap", "figures/TNF_alpha_Expression_Heatmap_top10.pdf", do_scale = TRUE, n_samples = 10, height = 10, width = 8)

generate_heatmap(TGF, invasive_log, invasive_pheno, "TGF-Beta Expression Heatmap", "figures/TGF_Beta_Expression_Heatmap_top10.pdf", do_scale = TRUE, n_samples = 10, height = 10, width = 8)
```

# check the batch2 normal vs MIS
```{r}
b2_pheno_merged
# set the comprison group
comp1 <- c("N", "P")
b2_comp1_pheno <- b2_pheno_merged[b2_pheno_merged$histopathology %in% comp1, ]

table(b2_comp1_pheno$histopathology)

# now subset the rawcount
comp1_raw <- tumor_raw_count_b2[, rownames(b2_comp1_pheno)]

b2_comp1_pheno <- b2_comp1_pheno[match(colnames(comp1_raw), rownames(b2_comp1_pheno)), ]

all(colnames(comp1_raw) == rownames(b2_comp1_pheno))  # Should return TRUE

# now creating DEG object 
comp1_dds <-  DESeqDataSetFromMatrix(countData = round(comp1_raw),
                                 colData = b2_comp1_pheno,
                                 design = ~ histopathology)

comp1_dds <- DESeq(comp1_dds)

comp1_vsd <- vst(comp1_dds, blind = FALSE)
# comparison
comp1_res <- results(comp1_dds, contrast = c("histopathology", "N", "P"))
comp1_res <- comp1_res[order(comp1_res$padj), ]
#comp1_significant_genes <- rownames(comp1_res[which(comp1_res$padj <= 0.05) & comp1_res[which(comp1_res$log2FoldChange >= 1.5), ])

```


```{r, fig.height= 8, fig.width = 10}
library(EnhancedVolcano)

# Specify the filename for the SVG file
output_filename <- "/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/B2_NvsP_DGEplot.pdf"
 
# Start SVG device
pdf(output_filename, width = 7, height = 8) # Set appropriate width and height

EnhancedVolcano(comp1_res,
    lab = rownames(comp1_res),
    title = 'Normal vs Precusor',
    #selectLab = c('PMEL','CDK2', 
     #             "CDHR1", "CDH1", "ITGA4"),
    x = 'log2FoldChange',
    y = 'pvalue',
    labSize = 5.0, 
    pointSize = 3.0, 
    FCcutoff = 1,
    pCutoff = 0.05,
    subtitle = bquote('FDR cut off 0.001 and absolute LFC cut off 1'))

dev.off()
```

# check the expression of MART1 in the invasive region

# read the stage 2 data
```{r}
stage2_tumor_pheno <- read.csv("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/processed_data/stageII_tumor_GeoMX_cluster_with_ROIs.csv")
table(stage2_tumor_pheno$cluster)
rownames(stage2_tumor_pheno) <- stage2_tumor_pheno$X
stage2_tumor_pheno$X <- NULL
stage2_tumor_pheno

# Assuming xy_locations is your data frame
ggplot(stage2_tumor_pheno, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(size = 4) +  # Points for each sample
  #geom_text(vjust = -0.5, hjust = 0.5, size = 3) +  # Labels for each point
  theme_minimal() +
  labs(title = "Correlation Spanning Tree - Sample Locations",
       x = "PC1",
       y = "PC2") +
  #scale_x_continuous(breaks = seq(floor(min(xy_locations$x)), ceiling(max(xy_locations$x)), by = 0.5)) +
  #scale_y_continuous(breaks = seq(floor(min(xy_locations$y)), ceiling(max(xy_locations$y)), by = 0.5)) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )

# now export the raw data
stage2_rawcount <- read.csv("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/processed_data/PCAII_merged_raw_05212024.csv")
rownames(stage2_rawcount) <- stage2_rawcount$X
stage2_rawcount$X <- NULL

# Replace periods with dashes except before ".dcc"
colnames(stage2_rawcount) <- gsub("\\.(?=.*\\.dcc)", "-", colnames(stage2_rawcount), perl = TRUE)

stage2_tumor_rawcount <- stage2_rawcount[, rownames(stage2_tumor_pheno)]
stage2_tumor_rawcount
```


```{r}
# now generate the dds object
dds <- DESeqDataSetFromMatrix(countData = round(stage2_tumor_rawcount),
                              colData = stage2_tumor_pheno, 
                              design = ~ cluster + batch)
dds <- DESeq(dds)

vsd <- vst(dds, blind = FALSE)

```


```{r}
dds <- estimateSizeFactors(dds)

norm_counts <- as.data.frame(counts(dds, normalized = TRUE))
norm_counts_matrix <- as.matrix(norm_counts)

#branch_gsva <- gsva(ssgseaParam(norm_counts_matrix, gene_signatures), verbose = F)
#branch_gsvaDF <- as.data.frame(branch_gsva)

```
# plot the HE result into PCA plot

```{r}
stage2_gsva <- gsva(ssgseaParam(norm_counts_matrix, selected_genes_converted_symbols), verbose = F)

stage2_gsvaDF <- as.data.frame(t(stage2_gsva))

stage2_gsvaDF

stage2_gsz <- as.data.frame(scale(stage2_gsvaDF))

```
```{r}
stage2_tumor_pheno$cluster <- as.character(stage2_tumor_pheno$cluster)
stage2_tumor_pheno
```

```{r figure 2, fig.height= 6, fig.width= 12}
breaks_list <- seq(-2, 2, by = 0.05)
print(length(breaks_list))
color_scale <- colorRampPalette(c("navy", "white", "red"))(length(breaks_list))

# Start SVG device
pdf("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/stage2_signature_analysis_heatmap.pdf", width = 12, height = 6)

cl_cb <- function(hcl, mat){
    # Recalculate manhattan distances for reorder method
    dists <- dist(stage2_gsz, method = "manhattan")

    # Perform reordering according to OLO method
    hclust_olo <- reorder(hcl, dists)
    return(hclust_olo)
}

# Define the colors for each category
hist_colors <- c(
  "epithelioid" = "#FF6347",            # Red
  "epithelioid_pleomorphism" = "#4682B4", # Blue
  "epithelioid_rhabdoid" = "#32CD32",    # Green
  "epithelioid_spindle" = "#FFD700",     # Yellow
  "pleomorphism" = "#DA70D6",            # Orchid
  "spindle" = "#8B4513",                 # SaddleBrown
  "stroma" = "#2F4F4F"                   # DarkSlateGray
)

# Define the colors for each cluster
cluster_colors <- c(
  "1" = "#FF6347",   # Red
  "2" = "#4682B4",   # Blue
  "3" = "#32CD32",   # Green
  "4" = "#FFD700"    # Yellow
)

pheatmap(t(stage2_gsz),
         scale = "row",
         #border_color = "grey",
         show_colnames = F,
         cellheight = 10,
         cellwidth = 1,
         treeheight_row = 15, 
         treeheight_col = 25,
         color = color_scale, 
         breaks = breaks_list, 
         cluster_cols = TRUE, 
         cluster_rows = FALSE,
         clustering_callback = cl_cb,
         annotation_col= stage2_tumor_pheno[, c("combined_hist","cluster")],
         annotation_colors = list(combined_hist = hist_colors,
                                  cluster = cluster_colors), 
          #                        case = mel_colors),
         fontsize = 12,         # General font size
         fontsize_row = 10,     # Font size for row names
         fontsize_col = 10,   
         legend = T,
         annotation_legend = TRUE)


# Manually render the heatmap using grid.draw
#grid::grid.draw(pheatmap_result$gtable)

# Close the PDF device
dev.off()
```


now ploting the batch1
```{r}
colnames(stage2_tumor_pheno)
```






```{r}
uni_features <- read.csv("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/HE_analysis/uni_mass100k-for-tk.csv")

uni_features[, c("patch.filename", "local.filepath", "slide", "histopathology", "umap.1", "umap.2","ROI.ID")]

# columns to remove
remove_cols <- c("patch.filename", "local.filepath", "slide", "histopathology", "umap.1", "umap.2","ROI.ID")

rownames(uni_features) <-  uni_features$patch.filename

uni_features_clean <- uni_features %>% 
 select(-all_of(remove_cols))

head(uni_features_clean)

# now perform PCA analysis
pca_result <- prcomp(uni_features_clean, scale. = TRUE)
pca_result
```
# now plot the PCA result
```{r}
explained_variance <- summary(pca_result)$importance[2, 1:2]
explained_variance

percent_pc1 <- round(explained_variance[1] * 100, 2)
percent_pc2 <- round(explained_variance[2] * 100, 2)

# Create a dataframe for plotting PCA results
pca_df <- as.data.frame(pca_result$x)

# Plot PC1 vs PC2 with explained variance percentages
# Make sure that the row names of 'uni_features' match the row names in 'pca_df'
pca_df$histopathology <- uni_features$histopathology
pca_df$slide <- uni_features$slide

# Plot PC1 vs PC2, colored by 'histopathology'
ggplot(pca_df, aes(x = PC1, y = PC2, color = slide)) +
  geom_point(size = 1) +
  labs(
    x = paste0("PC1 (", round(summary(pca_result)$importance[2, 1] * 100, 2), "% variance)"),
    y = paste0("PC2 (", round(summary(pca_result)$importance[2, 2] * 100, 2), "% variance)"),
    title = "PCA Plot (Colored by Histopathology)"
  ) +
  theme_minimal() +
  theme(legend.title = element_text(size = 12), legend.text = element_text(size = 10)) +
  scale_color_brewer(palette = "Set1")  # You can choose a different color palette if preferred
```

```{r}
library(plotly)
# Create a 3D scatter plot with plotly
p <- plot_ly(
  data = pca_df,
  x = ~PC1, y = ~PC2, z = ~PC3, 
  color = ~slide, 
  colors = "Set1",  # You can choose a different color palette if preferred
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 2)
) %>%
  layout(
    title = "3D PCA Plot (Colored by Histopathology)",
    scene = list(
      xaxis = list(title = paste0("PC1 (", round(summary(pca_result)$importance[2, 1] * 100, 2), "% variance)")),
      yaxis = list(title = paste0("PC2 (", round(summary(pca_result)$importance[2, 2] * 100, 2), "% variance)")),
      zaxis = list(title = paste0("PC3 (", round(summary(pca_result)$importance[2, 3] * 100, 2), "% variance)"))
    )
  )
p
```

```{r}
pca_df <- as.data.frame(pca_result$x)

# Select the top 30 PCs
top_30_pcs <- pca_df[, 1:30]

# Melt the PCA data for easier plotting
pca_melted <- melt(top_30_pcs)

# Plot the distribution of the top 30 principal components
ggplot(pca_melted, aes(x = value)) +
  geom_histogram(bins = 30, fill = "blue", color = "black", alpha = 0.7) +
  facet_wrap(~ variable, scales = "free") +  # Create a separate histogram for each PC
  labs(
    title = "Distribution of Top 30 Principal Components",
    x = "PC Value",
    y = "Frequency"
  ) +
  theme_minimal()
```
# read the original annotation file
```{r}
# Get the explained variance percentages for each PC
explained_variance <- summary(pca_result)$importance[2, ]

# Create a dataframe for the top 30 PCs and their explained variance
explained_variance_df <- data.frame(
  PC = paste0("PC", 1:length(explained_variance)),
  VarianceExplained = explained_variance * 100  # Convert to percentage
)

# Select the top 30 PCs
top_30_variance <- explained_variance_df[1:30, ]

# Plot the explained variance for the top 30 PCs
ggplot(top_30_variance, aes(x = PC, y = VarianceExplained)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(
    title = "Explained Variance of Top 30 Principal Components",
    x = "Principal Component",
    y = "Percentage of Variance Explained"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability
```


```{r, fig.height= 15, fig.width= 25}
breaks_list <- seq(-2, 2, by = 0.05)
print(length(breaks_list))
color_scale <- colorRampPalette(c("navy", "white", "red"))(length(breaks_list))


#output_filename <- paste0("/Users/yshi797/Desktop/PCA/PCAfigures_prep_TK/PCAIItumor_lineage_Invasive_heatmap_050924.pdf")
# Start SVG device
#pdf(output_filename, width = 50, height = 25) # Set appropriate width and height



pheatmap_result <- pheatmap(branch_gsvaDF,
         scale = "row",
         border_color = "grey",
         show_colnames = F,
         cellheight = 10,
         cellwidth = 3,
         treeheight_row = 0.5, 
         treeheight_col = 0.5,
         color = color_scale, 
         breaks = breaks_list, 
         annotation_col= stage2_tumor_pheno[, c(#"localization", 
                                     "histopathology", 
                                     #"MISvsInvasive", "Immune_TV", 
                                     "slide_name", "cluster")],
         #annotation_colors = list(histopathology = custom_colors, 
          #                        case = mel_colors),
         fontsize = 12,         # General font size
         fontsize_row = 10,     # Font size for row names
         fontsize_col = 10,   
         legend = T)
#dev.off()
print(pheatmap_result)


```

```{r}
branch_gsvaDFt <- as.data.frame(t(branch_gsvaDF))
branch_gsvaDFt["sample"] <- rownames(branch_gsvaDFt)
branch_gsvaDFt

stage2_tumor_pheno["sample"] <- rownames(stage2_tumor_pheno)

merged_data <- merge(branch_gsvaDFt, stage2_tumor_pheno, by = "sample")

merged_data
```

```{r}
merged_data$cluster <- as.character(merged_data$cluster)
merged_data$BRANCH1_SIGNATURE <- as.numeric(merged_data$BRANCH1_SIGNATURE)
merged_data$BRANCH3_SIGNATURE <- as.numeric(merged_data$BRANCH3_SIGNATURE)


summary(merged_data$cluster)
summary(merged_data$BRANCH1_SIGNATURE)
```


```{r comparison, fig.height= 6, fig.width= 8}
ggbetweenstats(data = merged_data,
               x = cluster,
               y = BRANCH1_SIGNATURE,
               messages = F,
               bf.message = F,
               mean.size = 2,
               mean.color = "black",
               ylab = "branch1 signature",
               pairwise.comparisons = FALSE,
               results.subtile = F,
               centrality.label.args = list(size = 4),
               ggsignif.args = list(textsize = 4, tip_length = 0.01)) + 
                theme(axis.title.y.right = element_blank(), 
                          axis.text.y.right = element_blank(), 
                          axis.ticks.y.right = element_blank(),
                          text = element_text(size = 12),       # Increase overall font size
                          axis.title = element_text(size = 12), # Increase axis titles font size
                          axis.text = element_text(size = 12))
                          #geom_signif(comparisons = list(c("Tumor", "Immune")), test.args = list(exact = FALSE)    
```

```{r branch3, fig.height= 6, fig.width= 8}
ggbetweenstats(data = merged_data,
               x = cluster,
               y = BRANCH3_SIGNATURE,
               messages = F,
               bf.message = F,
               mean.size = 2,
               mean.color = "black",
               ylab = "branch1 signature",
               pairwise.comparisons = FALSE,
               results.subtile = F,
               centrality.label.args = list(size = 4),
               ggsignif.args = list(textsize = 4, tip_length = 0.01)) + 
                theme(axis.title.y.right = element_blank(), 
                          axis.text.y.right = element_blank(), 
                          axis.ticks.y.right = element_blank(),
                          text = element_text(size = 12),       # Increase overall font size
                          axis.title = element_text(size = 12), # Increase axis titles font size
                          axis.text = element_text(size = 12))
                          #geom_signif(comparisons = list(c("Tumor", "Immune")), test.args = list(exact = FALSE)    
```

```{r}
# Assuming xy_locations is your data frame
ggplot(merged_data, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(size = 4) +  # Points for each sample
  #geom_text(vjust = -0.5, hjust = 0.5, size = 3) +  # Labels for each point
  theme_minimal() +
  labs(title = "PCA - Sample Locations",
       x = "PC1",
       y = "PC2") +
  #scale_x_continuous(breaks = seq(floor(min(xy_locations$x)), ceiling(max(xy_locations$x)), by = 0.5)) +
  #scale_y_continuous(breaks = seq(floor(min(xy_locations$y)), ceiling(max(xy_locations$y)), by = 0.5)) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )

ggplot(merged_data, aes(x = PC1, y = PC2, color = histopathology)) +
  geom_point(size = 4) +  # Points for each sample
  #geom_text(vjust = -0.5, hjust = 0.5, size = 3) +  # Labels for each point
  theme_minimal() +
  labs(title = "PCA - Sample Locations",
       x = "PC1",
       y = "PC2") +
  #scale_x_continuous(breaks = seq(floor(min(xy_locations$x)), ceiling(max(xy_locations$x)), by = 0.5)) +
  #scale_y_continuous(breaks = seq(floor(min(xy_locations$y)), ceiling(max(xy_locations$y)), by = 0.5)) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )

```

#load ovexpression spot, in total there are 11 overexpression spots
```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
library(msigdbr)
library(ReactomePA)
library(DOSE)

# List of spots (A to Q)
spots <- LETTERS[1:17]  # Generates letters from A to Q

# Directory paths
base_dir <- "PCAI_batch2_tumor_30x30_09082024_Results/CSV Sheets/Spot Lists/"

# Get Hallmark gene sets from MSigDB
hallmark_gene_sets <- msigdbr(species = "Homo sapiens", category = "H")

for (spot in spots) {
  # Read the CSV file for the current spot
  file_path <- paste0(base_dir, "Overexpression Spots ", spot, ".csv")
  spot_data <- read.csv(file_path)
  spot_genelist <- spot_data$X
  
  # Convert gene symbols to Entrez IDs
  entrez_ids <- bitr(spot_genelist, fromType = "SYMBOL", 
                     toType = "ENTREZID", OrgDb = org.Hs.eg.db)$ENTREZID
  
  # Check if entrez_ids is not empty
  if (length(entrez_ids) > 0) {
    
    ## 1. Perform GO enrichment analysis
    go_enrichment <- enrichGO(gene = entrez_ids, OrgDb = org.Hs.eg.db,
                              ont = "BP", pAdjustMethod = "BH", pvalueCutoff = 0.05)
    
    if (!is.null(go_enrichment) && nrow(go_enrichment) > 0) {
      go_plot <- dotplot(go_enrichment) + ggtitle(paste0("GO Biological Process Enrichment \n Spot ", spot, " vs others"))
      ggsave(paste0("Plots/GO_Enrichment_Spot_", spot, ".pdf"), plot = go_plot, width = 5, height = 6, dpi = 300)
    }
    
    ## 2. Perform KEGG enrichment analysis
    kegg_enrichment <- enrichKEGG(gene = entrez_ids, organism = "hsa",
                                  pAdjustMethod = "BH", pvalueCutoff = 0.05)
    
    if (!is.null(kegg_enrichment) && nrow(kegg_enrichment) > 0) {
      kegg_plot <- dotplot(kegg_enrichment) + ggtitle(paste0("KEGG Pathway Enrichment \n Spot ", spot, " vs others"))
      ggsave(paste0("Plots/KEGG_Enrichment_Spot_", spot, ".pdf"), plot = kegg_plot, width = 5, height = 6, dpi = 300)
    }
    
    ## 3. Perform Hallmark enrichment analysis
    hallmark_enrichment <- enricher(entrez_ids, TERM2GENE = hallmark_gene_sets[, c("gs_name", "entrez_gene")], 
                                    pAdjustMethod = "BH", pvalueCutoff = 0.05)
    
    if (!is.null(hallmark_enrichment) && nrow(hallmark_enrichment) > 0) {
      hallmark_plot <- dotplot(hallmark_enrichment) + ggtitle(paste0("Hallmark Gene Set Enrichment \n Spot ", spot, " vs others"))
      ggsave(paste0("Plots/Hallmark_Enrichment_Spot_", spot, ".pdf"), plot = hallmark_plot, width = 5, height = 6, dpi = 300)
    }
    
    ## 4. Perform Reactome enrichment analysis
    reactome_enrichment <- enrichPathway(gene = entrez_ids, organism = "human", 
                                         pAdjustMethod = "BH", pvalueCutoff = 0.05)
    
    if (!is.null(reactome_enrichment) && nrow(reactome_enrichment) > 0) {
      reactome_plot <- dotplot(reactome_enrichment) + ggtitle(paste0("Reactome Pathway Enrichment \n Spot ", spot, " vs others"))
      ggsave(paste0("Plots/Reactome_Enrichment_Spot_", spot, ".pdf"), plot = reactome_plot, width = 5, height = 6, dpi = 300)
    }
    
    ## 5. Perform DisGeNET enrichment analysis
    disgenet_enrichment <- enrichDGN(entrez_ids, pAdjustMethod = "BH", pvalueCutoff = 0.05)
    
    if (!is.null(disgenet_enrichment) && nrow(disgenet_enrichment) > 0) {
      disgenet_plot <- dotplot(disgenet_enrichment) + ggtitle(paste0("DisGeNET Disease Enrichment \n Spot ", spot, " vs others"))
      ggsave(paste0("Plots/DisGeNET_Enrichment_Spot_", spot, ".pdf"), plot = disgenet_plot, width = 5, height = 6, dpi = 300)
    }
    
  } else {
    message(paste0("No valid Entrez IDs for Spot ", spot))
  }
}

```



```{r}
go_enrichment <- enricher(gene = significant_genes, TERM2GENE = TERM2GENE, 
                          #pAdjustMethod = "BH", 
                          pvalueCutoff = 0.05)

dotplot(go_enrichment) + ggtitle("Hallmark Pathway Enrichment")
```

```{r plot the pathway enrichment, fig.height= 6, fig.width= 6}
# Visualize GO enrichment results
dotplot(go_enrichment) + ggtitle("GO Biological Process Enrichment \n branch1 vs branch3")

# Visualize KEGG enrichment results
dotplot(kegg_enrichment) + ggtitle("KEGG Pathway Enrichment \n branch1 vs branch3")
```

```{r prepare rank_list function}
# formats the ranked list for the fgsea() function
prepare_ranked_list <- function(ranked_list) { 
  # if duplicate gene names present, average the values
  if( sum(duplicated(ranked_list$Gene.name)) > 0) {
    ranked_list <- aggregate(.~Gene.name, FUN = mean, data = ranked_list)
    ranked_list <- ranked_list[order(ranked_list$log2FoldChange, decreasing = T),]
  }
  # omit rows with NA values
  ranked_list <- na.omit(ranked_list)
  # turn the dataframe into a named vector
  ranked_list <- tibble::deframe(ranked_list)
  ranked_list
}
```


```{r}
b2_pheno_merged
```


# show the invasive branch
```{r, fig.height= 6, fig.width= 6.5}
# Assuming xy_locations is your data frame
branch_plot <- ggplot(b2_pheno_merged, aes(x = spanning_tree_x, y = spanning_tree_y, fill = branch)) +
  geom_point(shape = 21, color = "black", size = 4, stroke = 1.2) +  # Points for each sample with black border
  theme_minimal() +
  labs(title = "Correlation Spanning Tree - Sample Locations",
       x = "X Coordinate",
       y = "Y Coordinate") +
  scale_fill_manual(values = c("branch1" = "tomato", "branch2" = "steelblue", "branch3" = "orange", "others" = "grey")) +  # Ensure "others" matches the case
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )

ggsave("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/branch_plots.pdf", 
       plot = branch_plot, width = 6.5, height = 6 , units = "in", device = "pdf", dpi = 300)
```


```{r, fig.height= 8, fig.width = 10}
library(EnhancedVolcano)

# Specify the filename for the SVG file
output_filename <- "/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/branch_DGEplot.pdf"
 
# Start SVG device
pdf(output_filename, width = 7, height = 8) # Set appropriate width and height

EnhancedVolcano(res,
    lab = rownames(res),
    selectLab = c("PCSK2", "CCL18", "TIMP1", "PRRX1", "S100A4", "S100B", "MPZ", 
                  #"CTSB", 
                  "A2M", "CXCL9", "HLA-A", "HLA-B", "HLA-C", "HLA-DR"),
    title = 'branch1 vs branch3',
    x = 'log2FoldChange',
    y = 'pvalue',
    labSize = 5.0, 
    pointSize = 3.0, 
    FCcutoff = 1.5,
    pCutoff = 0.001,
    subtitle = bquote('FDR cut off 0.001 and absolute LFC cut off 1'))

dev.off()
```
#generate heatmap for top and bottom 20 genes 
```{r}
resDF <- as.data.frame(res)

sig_res <- resDF %>% 
  filter(!is.na(padj)) %>% 
  filter(padj <= 0.05) %>% 
  arrange(desc(log2FoldChange))
  
sig_res

up_genes <- rownames(sig_res)[1:20]
dn_genes <- rownames(tail(sig_res, 20))
up_genes
```
```{r}
# now create dds project
b2_invasive_norm <- counts(invasive_dds, normalize = T)
b2_invasive_log <- log2(b2_invasive_norm + 1)
```



```{r show heatmap of the genes to make sure the order is right, fig.height= 12, fig.width = 10}
GOI <- c(up_genes, dn_genes)

#output_filename <- "/Users/yshi797/Desktop/PCA/PCAfigures_prep/morphology_DGEheatmap.pdf"
#pdf(output_filename, width = 4, height = 8) # Set appropriate width and height
pheatmap(b2_invasive_log[GOI, ], 
         scale = "row", 
         show_rownames = T, show_colnames = F, 
         cellheight = 12,
         cellwidth =3.5,
         #border_color = "black",
         treeheight_row = 1,
         treeheight_col = 1,
         cluster_rows = T,
         breaks = seq(-3, 3, 0.05),
         color = colorRampPalette(c("blue", "white", "red"))(120),
         annotation_col = invasive_pheno[, c("branch", "localization", "case")]) 
        # annotation_colors = list(case = anno_colors,
         #                         localization = loc_colors,
          #                        condition = condition_colors),


```

```{r}
b2_invasive_log_matrix <-  as.matrix(b2_invasive_log)
overall_gsva <- as.data.frame(t(gsva(ssgseaParam(b2_invasive_log_matrix, Overall_geneSig), verbose = F)))
overall_gsva <- overall_gsva %>% 
  rename(G1S = "CELL.CYCLE..G1.S", G2M = "CELL.CYCLE..G2.M")

```
```{r}
invasive_pheno_sig <- cbind(invasive_pheno, overall_gsva)
colnames(invasive_pheno_sig)
```


```{r genomic feature comparison, fig.height= 4, fig.width= 3.5}
set.seed(123)
emt_con <- ggbetweenstats(data = invasive_pheno_sig,
               x = branch,
               y = HALLMARK_G2M_CHECKPOINT,
               messages = F,
               bf.message = F,
               mean.size = 2,
               mean.color = "black",
               ylab = "HALLMARK_G2M_CHECKPOINT",
               pairwise.comparisons = FALSE,
               results.subtile = F,
               centrality.label.args = list(size = 4),
               ggsignif.args = list(textsize = 4, tip_length = 0.01)) + 
                theme(axis.title.y.right = element_blank(), 
                          axis.text.y.right = element_blank(), 
                          axis.ticks.y.right = element_blank(),
                          text = element_text(size = 12),       # Increase overall font size
                          axis.title = element_text(size = 12), # Increase axis titles font size
                          axis.text = element_text(size = 12))

print(emt_con)

```
```{r}
# Define the directory to save the plots
save_dir <- "/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/"

# Get the list of selected columns
selected_cols <- colnames(overall_gsva)

# Function to generate and save the plot for each column
generate_and_save_plot <- function(col_name, data, save_dir) {
  
  # Create the plot
  p <- ggbetweenstats(
    data = invasive_pheno_sig,
    x = branch,
    y = !!sym(col_name),  # Dynamically reference the column name
    messages = F,
    bf.message = F,
    mean.size = 2,
    mean.color = "black",
    ylab = col_name,  # Use the column name as y-label
    pairwise.comparisons = FALSE,
    results.subtitle = F,
    centrality.label.args = list(size = 4),
    ggsignif.args = list(textsize = 4, tip_length = 0.01)
  ) + 
    theme(
      axis.title.y.right = element_blank(), 
      axis.text.y.right = element_blank(), 
      axis.ticks.y.right = element_blank(),
      text = element_text(size = 12),       # Increase overall font size
      axis.title = element_text(size = 12), # Increase axis titles font size
      axis.text = element_text(size = 12)
    )
  
  # Define the file name
  file_name <- paste0(save_dir, col_name, ".pdf")
  
  # Save the plot
  ggsave(file_name, plot = p, width = 3.5, height = 4, units = "in", dpi = 300)
}

# Iterate through all selected columns, generate plots, and save
purrr::walk(selected_cols, ~generate_and_save_plot(.x, invasive_pheno_sig, save_dir))
```





# now try hallmark pathways
```{r hallmark, fig.height = 4, fig.width = 8}
library(stringr)
library(fgsea)
# need to do a filterating step before ranking
rank_res <- as.data.frame(res[order(res$log2FoldChange, decreasing=TRUE ), ])
rank_list <- rank_res[c("log2FoldChange")]

rank_list["Gene.name"] <- rownames(rank_list)
rank_list <- rank_list[c("Gene.name", "log2FoldChange")]

#filtered_rank_list <- na.omit(rank_list)

ranked_list <- prepare_ranked_list(rank_list)
head(ranked_list)

# generate GSEA results table using fgsea() by inputting the pathway list and ranked list
fgsea_results <- fgsea(pathways = hallmark_pathway_geneSigList,
                  stats = ranked_list,
                  minSize = 15,
                  maxSize = 500,
                  nperm= 1000)

fgsea_results %>% arrange (desc(NES)) %>% dplyr::select (pathway, padj, NES) %>% head()

waterfall_plot <- function(fgsea_results, graph_title, file_name) {
  plot <- fgsea_results %>%
    # filter(padj <= 0.05) %>% 
    # Remove 'HALLMARK_' prefix from the pathway title
    mutate(pathway = str_replace(pathway, "HALLMARK_", "")) %>%
    ggplot(aes(reorder(pathway, NES), NES)) +
      geom_bar(stat= "identity", aes(fill = padj <=  0.05), color = "black") +
      coord_flip() +
      labs(x = "Hallmark pathway", y = "Normalized Enrichment Score", title = graph_title) +
      scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "tomato")) +
      theme_bw() +
      theme(
        axis.text.y = element_text(size = 14),  # Change y-axis text size
        axis.text.x = element_text(size = 14),  # Change x-axis text size
        axis.title.y = element_text(size = 14),  # Change y-axis title size
        axis.title.x = element_text(size = 14),  # Change x-axis title size
        plot.title = element_text(size = 16, hjust = 1)  # Change plot title size
      )
  
  # Save the plot as an SVG file
  ggsave(file_name, plot = plot, width = 9 , height = 10, units = "in", device = "pdf", dpi = 300)
}

plot <- waterfall_plot(fgsea_results, "Hallmark pathway altered \n branch1 vs branch3", "/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/waterfall_branch1vsbranch3.pdf")

print(plot)
```
# generate the dotplot based on what we saw 

```{r}
selected_pathways <- c("HALLMARK_INTERFERON_ALPHA_RESPONSE", "HALLMARK_INTERFERON_GAMMA_RESPONSE", 
                       "HALLMARK_ANGIOGENESIS", "HALLMARK_HYPOXIA", "HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION", "HALLMARK_G2M_CHECKPOINT", 
                       "HALLMARK_TNFA_SIGNALING_VIA_NFKB", "HALLMARK_KRAS_SIGNALING_UP",
                       "HALLMARK_MYC_TARGETS_V1", "HALLMARK_INFLAMMATORY_RESPONSE")


# subset the data
subfgsea_res <- fgsea_results[fgsea_results$pathway %in% selected_pathways, ]

subfgsea_res$pathway <- gsub("HALLMARK_", "", subfgsea_res$pathway)

res_dots <- ggplot(subfgsea_res, aes(x = NES, y = reorder(pathway, NES), size = size, fill = padj)) +
  geom_point(shape = 21, color = "black", stroke = 1.5) +  # Adding black border
  scale_fill_gradient(low = "red", high = "steelblue", name = "Adjusted p-value") +  # Color fill based on padj
  scale_size_continuous(range = c(3, 10), name = "Gene set size") +
  labs(title = "Hallpark Pathway Enrichment \n Branch1 vs Branch3", x = "Normalized Enrichment Score (NES)", y = "Pathway") +
  theme_classic() + 
  theme(
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    panel.grid.major.y = element_line(color = "grey", linetype = "dashed"),  # Add dashed grey grid lines on y-axis
    panel.grid.major.x = element_line(color = "grey", linetype = "dashed")   # Add dashed grey grid lines on x-axis
  )

ggsave("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/branch1vsbranch3_pathway_enrichment_dotplot.pdf", plot = res_dots, width = 8, height = 6, dpi = 300)

```
# now generate breslow fisherexact plot


```{r breslow depth analysis}
selected_columns <- c("Codename", "depth..mm.")
breslow_anno <- read.csv("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/clinical_data/melanoma_transferred to LSP_3.0 09232022_DEIDENTIFIED_TV.csv")[, selected_columns]
breslow_anno

# Apply the function to create the new 'breslow_depth' column
# Define the function to assign breslow categories
breslow_anno$breslow_depth <- cut(
  breslow_anno$depth..mm., 
  breaks = c(-Inf, 1, 2, 4, Inf), 
  labels = c("<1mm", "1-2mm", "2-4mm", "MIS_only")
)

breslow_anno$breslow_depth <- as.character(breslow_anno$breslow_depth)

# View the result
# Ensure both key columns (Codename in breslow_anno and case in df) have the same name
names(breslow_anno)[names(breslow_anno) == "Codename"] <- "case"

# Perform the left join to add breslow_depth to df
merged_invasive_pheno <- invasive_pheno %>%
  left_join(breslow_anno[, c("case", "breslow_depth")], by = "case")

# manually assign MEL55
merged_invasive_pheno[merged_invasive_pheno$case_block == "MEL55 A2", "breslow_depth"] = "2-4mm"

# View the result
print(merged_invasive_pheno[, c("case", "breslow_depth")])
# subset <1mm and 1-2mm for fisther exact test
comparison <- c("<1mm", "1-2mm")
sub_breslow <- merged_invasive_pheno[merged_invasive_pheno$breslow_depth %in% comparison, c("case", "breslow_depth", "branch")]
table(sub_breslow$branch)

immune_pheno <-  merged_invasive_pheno[, c("case", "Immune_TV", "branch")]
table(immune_pheno$branch)
```

```{r}

write.csv(merged_invasive_pheno,
          file = paste0("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/PCAII_project_manuscript_preparation_2025/processed_data/", "Figure_5G.csv"),
          row.names = F)

```


```{r}
table(immune_pheno$Immune_TV)

# Create a contingency table
contingency_table <- table(merged_invasive_pheno$branch, merged_invasive_pheno$Immune_TV)

# Perform the Chi-squared test
chi_square_test_result <- chisq.test(contingency_table)

# View the result
chi_square_test_result

# Create a contingency table
contingency_table <- table(merged_invasive_pheno$branch, merged_invasive_pheno$Immune_TV)

# If small sample sizes, use Fisher's exact test
fisher_test_result <- fisher.test(contingency_table)

# View the result
fisher_test_result

p_value <- fisher_test_result$p.value

df_percentage <- immune_pheno %>%
  group_by(branch, Immune_TV) %>%
  summarise(count = n()) %>%
  mutate(percentage = count / sum(count) * 100)

df_percentage
 

immune_colors <- c("brisk TIL" = "red", "immune response" = "orange", "no immune" = "steelblue", "nonbrisk TIL" = "skyblue", "immune regression" = "darkblue")

# Create the bar plot with borders, custom colors, and Fisher's test annotation
immune_stackplot <- ggplot(df_percentage, aes(x = branch, y = percentage, fill = Immune_TV)) +
  geom_bar(stat = "identity", position = "fill", width = 0.7, color = "black", size = 0.5) +  # Add black border to bars
  scale_y_continuous(labels = scales::percent_format()) +  # Convert y-axis to percentage
  scale_fill_manual(values = immune_colors) +  # Use custom colors for Breslow depth
  labs(x = "Branch", y = "Percentage", fill = "Breslow Depth") +
  theme_minimal() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1),
        axis.text.y = element_text(size = 12),
        axis.text.x = element_text(size = 12),
        panel.grid.major.y = element_line(color = "grey", linetype = "dashed"),  # Add dashed grey grid lines on y-axis
        panel.grid.major.x = element_line(color = "grey", linetype = "dashed")   # Add dashed grey grid lines on x-axis
        ) +  # Add box around the plot
  annotate("text", x = 1.5, y = 1.05, label = "****", size = 5, hjust = 0)  # Add p-value

ggsave("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/branch1vsbranch3_immune_stackplot.pdf", plot = immune_stackplot, width = 4, height = 4, dpi = 300)

print(immune_stackplot)
```
```{r}
table(df_percentage$Immune_TV)
```


```{r}
write.csv(sub_breslow,
          file = paste0("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/PCAII_project_manuscript_preparation_2025/processed_data/", "Figure_5E.csv"),
          row.names = F)
```



```{r, fig.height= 4, fig.width=4}
# Create a contingency table
contingency_table <- table(sub_breslow$breslow_depth, sub_breslow$branch)

# Perform chi-squared test
chi_test <- chisq.test(contingency_table)
print(chi_test)

# If expected counts are low, use Fisher's exact test instead
fisher_test <- fisher.test(contingency_table)
print(fisher_test)

p_value <- fisher_test$p.value

df_percentage <- sub_breslow %>%
  group_by(branch, breslow_depth) %>%
  summarise(count = n()) %>%
  mutate(percentage = count / sum(count) * 100)

# Define custom colors for Breslow depth categories
custom_colors <- c("<1mm" = "pink", "1-2mm" = "darkorange")

# Create the bar plot with borders, custom colors, and Fisher's test annotation
breslow_stackplot <- ggplot(df_percentage, aes(x = branch, y = percentage, fill = breslow_depth)) +
  geom_bar(stat = "identity", position = "fill", width = 0.7, color = "black", size = 0.5) +  # Add black border to bars
  scale_y_continuous(labels = scales::percent_format()) +  # Convert y-axis to percentage
  scale_fill_manual(values = custom_colors) +  # Use custom colors for Breslow depth
  labs(x = "Branch", y = "Percentage", fill = "Breslow Depth") +
  theme_minimal() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1),
        axis.text.y = element_text(size = 12),
        axis.text.x = element_text(size = 12),
        panel.grid.major.y = element_line(color = "grey", linetype = "dashed"),  # Add dashed grey grid lines on y-axis
        panel.grid.major.x = element_line(color = "grey", linetype = "dashed")   # Add dashed grey grid lines on x-axis
        ) +  # Add box around the plot
  annotate("text", x = 1.5, y = 1.05, label = "****", size = 5, hjust = 0)  # Add p-value

#ggsave("/Users/yshi797/Partners HealthCare Dropbox/Yingxiao Shi/Dave Liu Lab/Yingxiao_Shi(TK)/Project3_PCA/figures/branch1vsbranch3_breslow_plot.pdf", plot = breslow_stackplot, width = 4, height = 4, dpi = 300)

print(breslow_stackplot)

```
```{r}

```



now perform PCA analysis on batch2 invasive regions
```{r}
b2_invasive_pheno <- b2_pheno_merged[(b2_pheno_merged$histopathology %in% c("RGP", "VGP")) &
                                       b2_pheno_merged$composition_TV %in% c("tumor", "melanocyte"),]

table(b2_invasive_pheno$branch)
b2_invasive_rawcount <- tumor_raw_count_b2[, rownames(b2_invasive_pheno)]
b2_invasive_rawcount
```



# now perform dds and PCA
```{r}
b2_invasive_dds <- DESeqDataSetFromMatrix(countData = round(b2_invasive_rawcount), # data needs to have the same order
                              colData = b2_invasive_pheno,
                              design = ~ branch)


b2_invasive_dds <- DESeq(b2_invasive_dds)
b2_invasive_vsd <- vst(b2_invasive_dds, blind = TRUE)

pcaData <- plotPCA(b2_invasive_vsd, intgroup = c("localization", "Immune_TV", "case", "branch"), returnData = T)
percentVar <- round(100 * attr(pcaData, "percentVar"))
```



```{r, fig.height= 6, fig.width= 6}
comb_p <- ggplot(pcaData, aes(PC1, PC2, fill=branch)) +
  geom_point(shape = 21, size= 3, color = "black") +
  labs(x = paste0("PC1: ",percentVar[1],"% variance"),
       y = paste0("PC2: ",percentVar[2],"% variance"),
       title = "PCA Plot colored by branch") +
  #scale_fill_manual(values=custom_colors) +  # Add this line
  #ggrepel::geom_text_repel(aes(label = case), 
   #                        color = "black") +
  theme(size = element_text(size = 10)) + 
  theme_bw()

#ggsave("PCA/PCAfigures_prep/mergedPCA_Immune_TV.png", height = 4, width = 4.5)
print(comb_p)
```

```{r}
b2_log
```

# read the HE annotation file
```{r}
selected_genes = c("IRF1","NGFR", "MLANA", "SOX9", "SOX10")

irf1_exp <- as.data.frame(t(b2_invasive_log[selected_genes, ]))
#irf1_exp <- irf1_exp %>% 
 # rename(IRF1_log = `b2_log["IRF1", ]`)

irf1_exp$sample_label <- rownames(irf1_exp)
irf1_exp

irf1_exp <- merge(b2_pheno_merged[, c("sample_label", "histopathology", "branch")], irf1_exp, by = "sample_label")
irf1_exp_invasive <- irf1_exp[irf1_exp$branch %in% c("invasive_branch1", "invasive_branch3"), ]
irf1_exp_invasive
```


```{r}

library(ggplot2)
library(ggsignif)
library(ggpubr)
# List of genes to compare
genes <- c("IRF1", "NGFR", "MLANA", "SOX9", "SOX10")

# Loop through each gene, generate the plot, and save as PDF
for (gene in genes) {
  # Create the plot
  gene_plot <- ggplot(irf1_exp_invasive, aes_string(x = "branch", y = gene)) +
    geom_boxplot(aes(fill = branch), outlier.shape = NA) +
    geom_jitter(width = 0.2, size = 1, alpha = 0.6) +
    theme_minimal() +
    theme(
      text = element_text(size = 16),   # Increase overall text size
      axis.title = element_text(size = 16), # Increase axis titles font size
      axis.text = element_text(size = 14)  # Increase axis text size
    ) +
    labs(y = paste(gene, "log"), x = "Branch") +
    stat_compare_means(method = "wilcox.test", label = "p.signif", size = 6)  # Wilcoxon test with larger p-value labels
  
  # Save the plot as high-resolution PDF
  ggsave(filename = paste0(gene, "_comparison.pdf"), plot = gene_plot, device = "pdf", height = 5, width = 5, dpi = 300)
  print(gene_plot)
}
```



```{r, brish match, fig.height= 6, fig.width= 8}
# Assuming xy_locations is your data frame
ggplot(b2_pheno_merged, aes(x = spanning_tree_x, y = spanning_tree_y, color = Immune_TV)) +
  geom_point(size = 4) +  # Points for each sample
  #geom_text(vjust = -0.5, hjust = 0.5, size = 3) +  # Labels for each point
  theme_minimal() +
  labs(title = "Correlation Spanning Tree - Sample Locations",
       x = "X Coordinate",
       y = "Y Coordinate") +
  scale_x_continuous(breaks = seq(floor(min(xy_locations$x)), ceiling(max(xy_locations$x)), by = 0.5)) +
  scale_y_continuous(breaks = seq(floor(min(xy_locations$y)), ceiling(max(xy_locations$y)), by = 0.5)) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12)
  )
```









